---
title: "Demo of User-Defined `log_pointpred` for `brms`'s Poisson Regression"
author: "Tingxuan Wu and Longhai Li"
format: 
  html:
    number-sections: false
    code-fold: true
vignette: >
  %\VignetteIndexEntry{Demo of user-defined log_pointpred for Poisson via brms}
  %\VignetteEncoding{UTF-8}
  %\VignetteEngine{quarto::html}
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(
  fig.width = 7,
  fig.height = 5,
  fig.align = "center"
)
set.seed(13)

# Global toggle to force regeneration of heavy computations
force_rerun <- FALSE

# Ensure the resources directory exists to save heavy computations
if (!dir.exists("resources")) {
  dir.create("resources", recursive = TRUE)
}
```

```{r load-packages}
#| code-fold: show
#| message: false
#| warning: false
suppressWarnings(unloadNamespace("Zresidual"))
library(Zresidual)
library(brms)
library(knitr)
library(kableExtra)
```

## 1. Define a Custom Predictive Function

`Zresidual` is designed to be extensible. By defining a function named `log_pointpred_<backend>_<family>` in your Global Environment, `Zresidual` will automatically detect and use it instead of its internal defaults. 

For a `brms` Poisson model, the backend class is `brmsfit` and the family is `poisson`, so we name the function `log_pointpred_brmsfit_poisson`.

```{r}
#| eval: true 
#| label: custom-log-pointpred
#| code-fold: show
log_pointpred_brmsfit_poisson <- function(fit, data = NULL, ...) {
  
  # 1. Resolve data
  data_in <- if (is.null(data)) fit$data else data
  
  # 2. Extract observed response
  model.data <- model.frame(fit$formula, data = data_in)
  model.var  <- names(model.data)
  
  response <- fit$formula$resp
  if (is.null(response) || !response %in% model.var) {
    response <- all.vars(fit$formula$formula)[1]
  }
  
  sim.y <- as.numeric(model.data[[response]])
  n <- length(sim.y)
  
  # 3. Extract Posterior Expected Values (lambda)
  # brms::posterior_epred safely extracts the distribution parameter 'mu' (which is lambda for Poisson)
  lambda <- brms::posterior_epred(fit, dpar = "mu", newdata = data_in, ...)
  mc_used <- nrow(lambda)
  
  # 4. Calculate Log-PMF and Log-Survival Functions
  lpmf_hat <- matrix(NA_real_, mc_used, n)
  lsf_hat  <- matrix(NA_real_, mc_used, n)
  
  for (i in seq_len(n)) {
    lpmf_hat[, i] <- dpois(sim.y[i], lambda = lambda[, i], log = TRUE)
    lsf_hat[, i]  <- ppois(sim.y[i], lambda = lambda[, i],
                           lower.tail = FALSE, log.p = TRUE)
  }
  
  list(
    lpmf_hat = lpmf_hat,
    lsf_hat  = lsf_hat
  )
}
```

## 2. Simulating a Non-linear Count Dataset

We simulate a count dataset where the true log-rate ($\eta$) follows a sine wave over $x \in [0, 10]$. To avoid any potential matrix-parsing issues with inline functions in `brms` formulas, we explicitly compute and store `sinx` in the dataset.

```{r}
#| label: simulate-data
#| code-fold: show
n <- 200
x <- runif(n, 0, 10)
sinx <- sin(x)

# True non-linear relationship (sine wave for log-rate)
eta_true <- 1.5 * sin(x) 
y <- rpois(n, exp(eta_true))

dat <- data.frame(y = y, x = x, sinx = sinx)

cat("Mean count:", mean(dat$y), "\nMax count:", max(dat$y), "\n")
```

## 3. Fitting Models and Calculating Z-residuals

We fit a **Wrong Model** (linear) and a **Correct Model** (using the explicitly created `sinx` term). The fitted models are saved to disk with the `brms_pois_` prefix to prevent collision with other demos.

```{r}
#| label: fit-brms-models
#| results: hide
#| message: false
#| warning: false

refit_behavior <- ifelse(force_rerun, "always", "on_change")

# 1. Wrong Model (Linear)
fit_wrong <- brm(
  y ~ x, 
  family = poisson(), 
  data = dat,
  file = "resources/brms_pois_wrong", 
  file_refit = refit_behavior,
  refresh = 0,
  cores = 4
)

# 2. Correct Model (Non-linear using explicit sinx predictor)
fit_correct <- brm(
  y ~ sinx, 
  family = poisson(), 
  data = dat,
  file = "resources/brms_pois_correct", 
  file_refit = refit_behavior,
  refresh = 0,
  cores = 4
)
```

We evaluate the models using standard Posterior Predictive P-values (`method = "post"`). `Zresidual` will automatically detect `log_pointpred_brmsfit_poisson` from the Global Environment.

```{r}
#| label: calc-zresiduals
#| code-fold: show

# Calculate Z-residuals using the custom posterior predictive function
z_wrong   <- Zresidual(fit_wrong, data = dat, config = list(method = "post"), randomized = TRUE, nrep = 10,
log_pointpred = log_pointpred_brmsfit_poisson)
z_correct <- Zresidual(fit_correct, data = dat, config = list(method = "post"), randomized = TRUE, nrep = 10,
log_pointpred = log_pointpred_brmsfit_poisson)

# Extract Covariates for Diagnostics
info_wrong   <- Zcov(fit_wrong, data = dat)
info_correct <- Zcov(fit_correct, data = dat)
```

## 4. Diagnostics with Z-residuals

```{r}
#| label: diagnosis-comparison
#| code-fold: true
#| fig-width: 12
#| fig-height: 8

i <- 1 # Change this value (1 to 10) to see other randomized replicates

par(mfrow = c(2, 3), mar = c(4, 4, 3, 1))

# --- ROW 1: WRONG MODEL (Linear) ---
qqnorm(z_wrong, info = info_wrong, irep = i, main = "Wrong: Z-Resid Q-Q")

plot(z_wrong, info = info_wrong, x_axis_var = "x", 
     category = dat$y, irep = i, main = "Wrong: Z-Resid vs X")

boxplot(z_wrong, info = info_wrong, x_axis_var = "lp", 
        irep = i, main = "Wrong: Z-Resid vs LP")

# --- ROW 2: CORRECT MODEL (Explicit Non-linear) ---
qqnorm(z_correct, info = info_correct, irep = i, main = "Correct: Z-Resid Q-Q")

plot(z_correct, info = info_correct, x_axis_var = "x", 
     category = dat$y, irep = i, main = "Correct: Z-Resid vs X")

boxplot(z_correct, info = info_correct, x_axis_var = "lp", 
        irep = i, main = "Correct: Z-Resid vs LP")
```

**Replicated p-values of the same fitted models**
```{r}
#| label: replicated-p-values-spanned
#| code-fold: true

res_table <- data.frame(
  "Replicate" = paste("Rep", 1:ncol(z_wrong)),
  "SW_W"      = as.numeric(sw.test.zresid(z_wrong)),
  "AOV_W"     = as.numeric(aov.test.zresid(z_wrong, X = "lp", zcov = info_wrong)),
  "BL_W"      = as.numeric(bartlett.test.zresid(z_wrong, X = "lp", zcov = info_wrong)),
  "SW_C"      = as.numeric(sw.test.zresid(z_correct)),
  "AOV_C"     = as.numeric(aov.test.zresid(z_correct, X = "lp", zcov = info_correct)),
  "BL_C"      = as.numeric(bartlett.test.zresid(z_correct, X = "lp", zcov = info_correct))
)

res_table %>%
  kable(
    digits = 4, 
    col.names = c("Replicate", "SW", "AOV", "Bartlett", "SW", "AOV", "Bartlett"),
    caption = "Diagnostic Test p-values Across Randomized Replicates",
    align = "lcccccc"
  ) %>%
  kable_styling(
    bootstrap_options = c("striped", "condensed"), 
    full_width = FALSE,
    position = "center"
  ) %>%
  add_header_above(c(" " = 1, "Wrong Model (Linear)" = 3, "Correct Model (Non-linear)" = 3))
```

## 5. ANOVA Test Distributions (Simulation Loop)

We run a simulation loop to verify that our custom `log_pointpred` implementation correctly handles the explicitly generated `sinx` covariate and produces appropriate $p$-values.

```{r}
#| label: replication-verify-anova
#| code-fold: true
#| results: hide
#| message: false
#| warning: false
#| eval: false

n_sim <- 50
sim_file <- "resources/brms_pois_simulation_results.rds"

if (!force_rerun && file.exists(sim_file)) {
  sim_res <- readRDS(sim_file)
  p_c <- sim_res$p_c
  p_w <- sim_res$p_w
} else {
  p_c <- numeric(n_sim)
  p_w <- numeric(n_sim)
  n_sample <- 200 
  
  for(i in 1:n_sim) {
    # New independent data
    x_sim <- runif(n_sample, 0, 10)
    sinx_sim <- sin(x_sim)
    y_sim   <- rpois(n_sample, exp(1.5 * sinx_sim))
    dat_sim <- data.frame(y = y_sim, x = x_sim, sinx = sinx_sim)
    
    # --- Case 1: Wrong Model (Update Fit) ---
    # Using update() bypasses C++ compilation, reusing the compiled model 
    # from 'fit_wrong' and significantly speeding up the loop.
    fit_w_sim <- update(
      fit_wrong, 
      newdata = dat_sim,
      refresh = 0, 
      silent = 2,
      init = "0",          # Start from zero, not previous posterior
      iter = 2000,         # Ensure enough iterations for convergence
      warmup = 1000
    )
    
    z_w_sim <- Zresidual(fit_w_sim, data = dat_sim, config = list(method = "post"), randomized = TRUE, nrep = 1)
    info_w_sim <- Zcov(fit_w_sim, data = dat_sim)
    p_w[i] <- as.numeric(aov.test.zresid(z_w_sim, X = "lp", zcov = info_w_sim))
    
    # --- Case 2: Correct Model (Update Fit) ---
    fit_c_sim <- update(
      fit_correct, 
      newdata = dat_sim,
      refresh = 0, 
      silent = 2,
      init = "0",          # Independent start
      iter = 2000, 
      warmup = 1000
    )
    
    z_c_sim <- Zresidual(fit_c_sim, data = dat_sim, config = list(method = "post"), randomized = TRUE, nrep = 1)
    info_c_sim <- Zcov(fit_c_sim, data = dat_sim)
    p_c[i] <- as.numeric(aov.test.zresid(z_c_sim, X = "lp", zcov = info_c_sim))
  }
  
  saveRDS(list(p_c = p_c, p_w = p_w), sim_file)
}
```

```{r}
#| label: replication-hists
#| fig-width: 10
#| fig-height: 5
#| code-fold: true
#| eval: false

par(mfrow = c(1, 2))
hist(p_c, main = "Correct: Model P-values", xlab = "p-value", col = "lightblue", breaks = 10)
abline(h = n_sim / 10, col = "red", lty = 2)

hist(p_w, main = "Wrong: Small p-values", xlab = "p-value", col = "salmon", breaks = 10)
```

**Discussion:** The distribution of $p$-values for the correctly specified model confirms that our custom predictive function `log_pointpred_brmsfit_poisson` is functioning flawlessly. Using an explicit predictor variable rather than an inline formula computation ensures accurate posterior predictions. Conversely, the skewed distribution for the wrong model demonstrates the test's high power in detecting misspecification in count data.