---
title: "Demo of Z-residual Diagnosis for brms Logistic Regression"
author: "Tingxuan Wu and Longhai Li"
format: 
  html:
    number-sections: false
    code-fold: true
vignette: >
  %\VignetteIndexEntry{Demo of Z-residual diagnosis for logistic regression via brms}
  %\VignetteEncoding{UTF-8}
  %\VignetteEngine{quarto::html}
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(
  fig.width = 7,
  fig.height = 5,
  fig.align = "center"
)
set.seed(123)

# Global toggle to force regeneration of heavy computations
force_rerun <- FALSE

# Ensure the resources directory exists to save heavy computations
if (!dir.exists("resources")) {
  dir.create("resources", recursive = TRUE)
}
```

```{r load-packages}
#| code-fold: show
#| message: false
#| warning: false
suppressWarnings(unloadNamespace("Zresidual"))
library(Zresidual)
library(brms)
library(knitr)
library(kableExtra)
```

## 1. Simulating a Non-linear Dataset

Because `Zresidual` provides native support for models fitted via the `brms` package, we do not need to construct a custom predictive function. We can jump straight into simulating our dataset. 

We simulate a dataset where the true log-odds follow a sine wave. With $x \in [0, 10]$, the average log-odds are near zero, naturally keeping prevalence $\approx 0.5$.

```{r}
#| label: simulate-data
#| code-fold: show
n <- 200
x <- runif(n, 0, 10)

# True non-linear relationship (sine wave)
eta_true <- 2.5 * sin(x) 
y <- rbinom(n, 1, plogis(eta_true))

dat <- data.frame(y = y, x = x)

cat("Prevalence (proportion of 1s):", mean(dat$y), "\n")
```

## 2. Fitting Models and Calculating Z-residuals

We compare a **Wrong Model** (assuming a linear relationship) against a **Correct Model** (specifying the sine term) using Bayesian logistic regression via `brms`. By specifying the `file` argument, `brms` will save the fitted model. If `force_rerun` is `TRUE`, it uses `file_refit = "always"` to force a fresh MCMC run.

```{r}
#| label: fit-brms-models
#| results: hide
#| message: false
#| warning: false

# Determine refit behavior based on the global toggle
refit_behavior <- ifelse(force_rerun, "always", "on_change")

# 1. Wrong Model (Linear)
fit_wrong <- brm(
  y ~ x, 
  family = bernoulli(), 
  data = dat,
  file = "resources/brms_logit_wrong", 
  file_refit = refit_behavior,
  refresh = 0,
  cores = 4
)

# 2. Correct Model (Non-linear)
fit_correct <- brm(
  y ~ sin(x), 
  family = bernoulli(), 
  data = dat,
  file = "resources/brms_logit_correct", 
  file_refit = refit_behavior,
  refresh = 0,
  cores = 4
)
```

Next, we calculate the Z-residuals. For `brms` fits, `Zresidual` uses the Importance Sampling Cross-Validation (`iscv`) method by default. We also extract the covariate data using `Zcov()` to facilitate plotting.

```{r}
#| label: calc-zresiduals
#| code-fold: show

# Calculate Z-residuals (10 replicates each)
z_wrong   <- Zresidual(fit_wrong, data = dat, config = list(method = "iscv"), randomized = TRUE, nrep = 10)
z_correct <- Zresidual(fit_correct, data = dat, config = list(method = "iscv"), randomized = TRUE, nrep = 10)

# Extract Z-covariate information (No need for type = "zero" on standard logistic models)
info_wrong   <- Zcov(fit_wrong, data = dat)
info_correct <- Zcov(fit_correct, data = dat)
```

## 3. Diagnostics with Z-residuals

The plots below display the diagnostics for a single set of Z-residuals.

```{r}
#| label: diagnosis-comparison
#| code-fold: true
#| fig-width: 12
#| fig-height: 8

# Change this value (1 to 10) to see other replicated plots
i <- 1

# Set up 2 rows (Wrong vs Correct) and 3 columns (QQ, Scatter, Box)
par(mfrow = c(2, 3), mar = c(4, 4, 3, 1))

# --- ROW 1: WRONG MODEL (Linear) ---
# 1. Q-Q Plot
qqnorm(z_wrong, info = info_wrong, irep = i, main = "Wrong: Z-Resid Q-Q")

# 2. Scatter Plot (Note the wave pattern in the residuals, categorized by binary y)
plot(z_wrong, info = info_wrong, x_axis_var = "x", 
     category = dat$y, irep = i, main = "Wrong: Z-Resid vs X")

# 3. Boxplot
boxplot(z_wrong, info = info_wrong, x_axis_var = "lp", 
        irep = i, main = "Wrong: Z-Resid vs LP")

# --- ROW 2: CORRECT MODEL (Non-linear) ---
# 1. Q-Q Plot
qqnorm(z_correct, info = info_correct, irep = i, main = "Correct: Z-Resid Q-Q")

# 2. Scatter Plot (Residuals look like random noise, categorized by binary y)
plot(z_correct, info = info_correct, x_axis_var = "x", 
     category = dat$y, irep = i, main = "Correct: Z-Resid vs X")

# 3. Boxplot
boxplot(z_correct, info = info_correct, x_axis_var = "lp", 
        irep = i, main = "Correct: Z-Resid vs LP")
```

> **Note:** To inspect different realizations of the randomized residuals, change the value of `i` (the `irep` index) in the code chunk above.

**Replicated p-values of the same fitted models**
```{r}
#| label: replicated-p-values-spanned
#| code-fold: true

# 1. Create the data frame 
res_table <- data.frame(
  "Replicate" = paste("Rep", 1:ncol(z_wrong)),
  "SW_W"      = as.numeric(sw.test.zresid(z_wrong)),
  "AOV_W"     = as.numeric(aov.test.zresid(z_wrong, X = "lp", zcov = info_wrong)),
  "BL_W"      = as.numeric(bartlett.test.zresid(z_wrong, X = "lp", zcov = info_wrong)),
  "SW_C"      = as.numeric(sw.test.zresid(z_correct)),
  "AOV_C"     = as.numeric(aov.test.zresid(z_correct, X = "lp", zcov = info_correct)),
  "BL_C"      = as.numeric(bartlett.test.zresid(z_correct, X = "lp", zcov = info_correct))
)

# 2. Render with minimalistic styling and spanning headers
res_table %>%
  kable(
    digits = 4, 
    col.names = c("Replicate", "SW", "AOV", "Bartlett", "SW", "AOV", "Bartlett"),
    caption = "Diagnostic Test p-values Across Randomized Replicates",
    align = "lcccccc"
  ) %>%
  kable_styling(
    bootstrap_options = c("striped", "condensed"), 
    full_width = FALSE,
    position = "center"
  ) %>%
  add_header_above(c(" " = 1, "Wrong Model (Linear)" = 3, "Correct Model (Non-linear)" = 3))
```

## 4. ANOVA Test Distributions of Multiple Simulated Datasets 

We perform a simulation loop to compare the $p$-values produced by the ANOVA diagnostic test. 

*Note: Bayesian MCMC sampling is computationally heavy. To keep compilation times reasonable, we save the results of this loop to the `resources` directory and skip execution on subsequent builds unless `force_rerun` is enabled.*

```{r}
#| label: replication-verify-anova
#| code-fold: true
#| results: hide
#| message: false
#| warning: false

n_sim <- 50
sim_file <- "resources/brms_logit_simulation_results.rds"

if (!force_rerun && file.exists(sim_file)) {
  # Load pre-computed results if they exist and force_rerun is FALSE
  sim_res <- readRDS(sim_file)
  p_c <- sim_res$p_c
  p_w <- sim_res$p_w
} else {
  # Otherwise, execute the heavy simulation loop
  p_c <- numeric(n_sim)
  p_w <- numeric(n_sim)
  n_sample <- 200 
  
  for(i in 1:n_sim) {
    x_sim <- runif(n_sample, 0, 10)
    
    # Generate true non-linear data
    eta_sim <- 2.5 * sin(x_sim)
    y_sim   <- rbinom(n_sample, 1, plogis(eta_sim))
    dat_sim <- data.frame(y = y_sim, x = x_sim)
    
    # --- Case 1: Wrong Model (Linear) ---
    fit_w_sim <- update(fit_wrong, newdata = dat_sim, refresh = 0)
    z_w_sim   <- Zresidual(fit_w_sim, data = dat_sim, config = list(method = "iscv"), randomized = TRUE, nrep = 1)
    info_w_sim <- Zcov(fit_w_sim, data = dat_sim)
    
    p_w[i] <- as.numeric(aov.test.zresid(z_w_sim, X = "lp", zcov = info_w_sim))
    
    # --- Case 2: Correct Model (Sine) ---
    fit_c_sim <- update(fit_correct, newdata = dat_sim, refresh = 0)
    z_c_sim   <- Zresidual(fit_c_sim, data = dat_sim, config = list(method = "iscv"), randomized = TRUE, nrep = 1)
    info_c_sim <- Zcov(fit_c_sim, data = dat_sim)
    
    p_c[i] <- as.numeric(aov.test.zresid(z_c_sim, X = "lp", zcov = info_c_sim))
  }
  
  # Save the results for future renders
  saveRDS(list(p_c = p_c, p_w = p_w), sim_file)
}
```

```{r}
#| label: replication-hists
#| fig-width: 10
#| fig-height: 5
#| code-fold: true

par(mfrow = c(1, 2))
hist(p_c, main = "Correct: Uniform p-values", xlab = "p-value", col = "lightblue", breaks = 10)
abline(h = n_sim / 10, col = "red", lty = 2)

hist(p_w, main = "Wrong: Small p-values", xlab = "p-value", col = "salmon", breaks = 10)
```

**Discussion:** For the **Correct Model**, the distribution of $p$-values is relatively uniform, indicating a well-specified model. For the **Wrong Model**, most $p$-values are small (clustered near 0), indicating that the ANOVA test successfully detected the non-linear trend despite the Bayesian sampling constraints.