---
title: "Z-residual Diagnostic Tool for Assessing Covariate Functional Form in Shared Frailty Models"
date: today
format:
  html:
    toc: true
    number-sections: true
    format-links: false
    page-layout: full
    code-fold: true
    html-math-method: mathjax
execute:
  echo: true
  warning: false
  message: false
  cache: false
  freeze: false
crossref:
      eq-prefix: "Equation"
editor_options: 
  chunk_output_type: console
bibliography: ref.bib
link-citations: true      # for HTML
vignette: >
  %\VignetteIndexEntry{Z-residual diagnostic for shared frailty models}
  %\VignetteEncoding{UTF-8}
  %\VignetteEngine{knitr::rmarkdown}
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(
 fig.width = 7, fig.height = 5, fig.align = "center"
)
set.seed(123)

# Global control to force regeneration of GIFs
force_rerun <- FALSE 
```

```{r}
#| label: setup-paths
#| include: false

# 1. Identify Basename and Folder Name
current_file <- knitr::current_input()
if (is.null(current_file)) current_file <- "Zresidual_vignette.qmd" 

base_name <- tools::file_path_sans_ext(current_file)
folder_name <- paste0(base_name, "_savedfiles")

# 2. Configuration
pkg_name <- "Zresidual"
is_source <- file.exists("DESCRIPTION") || file.exists("../DESCRIPTION")
root_dir  <- ifelse(file.exists("DESCRIPTION"), ".", "..")

# 3. Define Paths
if (is_source) {
  # Development: Save permanently in inst/extdata/[basename]_savedfiles
  extdata_path <- file.path(root_dir, "inst", "extdata", folder_name)
  if (!dir.exists(extdata_path)) dir.create(extdata_path, recursive = TRUE)
  is_dev <- TRUE
} else {
  # Installed: Read from system files
  extdata_path <- system.file("extdata", folder_name, package = pkg_name)
  is_dev <- FALSE
}

# Local rendering folder (current directory)
local_assets_dir <- folder_name
if (!dir.exists(local_assets_dir)) dir.create(local_assets_dir, recursive = TRUE)

# 4. Helper: Local Asset Management
get_local_asset <- function(filename, source_dir, target_dir) {
  src <- file.path(source_dir, filename)
  dest <- file.path(target_dir, filename)
  
  if (!file.exists(src)) return(NULL)
  
  if (!file.exists(dest) || file.mtime(src) > file.mtime(dest)) {
    file.copy(src, dest, overwrite = TRUE)
  }
  return(dest)
}
```

## Installing Zresidual and Required Packages

### Installing Z-residual from Source (Developer Use)

```{r}
#| label: install-locally
#| eval: false
#| include: true
# For developer use to refresh the local installation
remove.packages("Zresidual")
devtools::document()
devtools::install()
```

### Installing Z-residual from GitHub

```{r}
#| label: install-and-load-zresidual
#| eval: false
if (!requireNamespace("Zresidual", quietly = TRUE)) {
  if (!requireNamespace("remotes", quietly = TRUE)) install.packages("remotes")
  remotes::install_github("tiw150/Zresidual", upgrade = "never", dependencies = TRUE)
}
```

```{r}
library(Zresidual)
```

### Loading R Packages used in this Demo

```{r loading-packages}
pkgs <- c(
  "survival", "EnvStats", "foreach", "statip", "VGAM", "plotrix", "actuar",
  "stringr", "Rlab", "dplyr", "rlang", "tidyr",
  "matrixStats", "timeDate", "katex", "gt", "loo"
)

missing_pkgs <- pkgs[!vapply(pkgs, requireNamespace, logical(1), quietly = TRUE)]
if (length(missing_pkgs)) {
  message("Installing missing packages: ", paste(missing_pkgs, collapse = ", "))
  install.packages(missing_pkgs, dependencies = TRUE)
}

invisible(lapply(pkgs, function(p) {
  suppressPackageStartupMessages(library(p, character.only = TRUE))
}))

nc <- parallel::detectCores(logical = FALSE)
if (!is.na(nc) && nc > 1) options(mc.cores = nc - 1)
```

## Introduction

This vignette explains how to use the `Zresidual` package to calculate Z-residuals based on the output of the `coxph` function from the `survival` package in R. It also demonstrates how Z-residuals can be used to assess the overall goodness of fit (GOF) and identify specific model misspecifications in semi-parametric shared frailty models. For a thorough understanding of the underlying theory, please refer to: *"Z-residual diagnostics for detecting misspecification of the functional form of covariates for shared frailty models."*

## Definition of Z-residual

We use Z-residuals to diagnose shared frailty models in a Cox proportional hazards setting where the baseline function is unspecified. For a group $i$ with $n_i$ individuals, let $y_{ij}$ be a possibly right-censored observation and $\delta_{ij}$ be the indicator for being uncensored. The normalized randomized survival probabilities (RSPs) are defined as:

$$
S_{ij}^{R}(y_{ij}, \delta_{ij}, U_{ij}) =
\left\{
\begin{array}{rl}
S_{ij}(y_{ij}), & \text{if } \delta_{ij}=1, \\
U_{ij}\,S_{ij}(y_{ij}), & \text{if } \delta_{ij}=0, 
\end{array}
\right.
$$ {#eq-rsp}

where $U_{ij} \sim \text{Uniform}(0, 1)$ and $S_{ij}(\cdot)$ is the postulated survival function. RSPs are transformed into Z-residuals via the normal quantile function:

$$
r_{ij}^{Z}(y_{ij}, \delta_{ij}, U_{ij})=-\Phi^{-1} (S_{ij}^R(y_{ij}, \delta_{ij}, U_{ij}))
$$ {#eq-zresid}


Under the true model, Z-residuals are normally distributed. This transformation allows researchers to leverage traditional normal-regression diagnostic tools for censored data. Furthermore, normal transformation highlights extreme RSPs that may indicate model misspecification but could otherwise be overlooked.

## Examples for Illustration and Demonstration

### Load the real Dataset

We utilize data from 411 acute myeloid leukemia patients recorded at the M. D. Anderson Cancer Center (1980–1996). The dataset focuses on patients under 60 from 24 districts. Key variables include survival time, age, sex, white blood cell count (WBC), and the Townsend deprivation score (TPI).

```{r data-read}
data_path <- system.file("extdata", "LeukSurv.rda", package = "Zresidual")
load(data_path)

LeukSurv <- transform(LeukSurv,
  district = as.factor(district),
  sex      = as.factor(sex),
  logwbc   = log(wbc + 0.001)
)
LeukSurv <- LeukSurv[LeukSurv$age < 60, ]
```

### Fitting Models

We compare two models: the **wbc model** (using raw WBC) and the **lwbc model** (using log-transformed WBC).

```{r fit-models}
fit_LeukSurv_wbc <- coxph(Surv(time, cens) ~ age + sex + wbc + tpi +
          frailty(district, distribution="gamma"), data = LeukSurv)

fit_LeukSurv_logwbc <- coxph(Surv(time, cens) ~ age + sex + logwbc + tpi +
          frailty(district, distribution="gamma"), data = LeukSurv)
```

### Computing Z-Residuals 

Once the model is fitted, we calculate Z-residuals for both models using 1,000 repetitions to account for randomization in censored cases.

```{r zresiduals-linear, cache=FALSE}
Zresid.LeukSurv.wbc <- Zresidual(object = fit_LeukSurv_wbc, nrep = 1000)
Zresid.LeukSurv.logwbc <- Zresidual(object = fit_LeukSurv_logwbc, nrep = 1000)
```

### Inspecting the Normality of Z-Residuals for Checking Overall GOF

Under a correctly specified shared frailty model, Z-residuals should be approximately standard normal, so a Q–Q plot should closely follow the 45° reference line. Because Z-residuals are generated from randomized survival probabilities for censored observations, the realized Z-residual set (and any normality-test p-value) can vary from one randomization replicate to another.

To assess overall goodness-of-fit (GOF), we primarily use Q–Q plots, and we report the replicate-specific Shapiro–Wilk (Z-SW) normality-test p-value as a compact numerical summary. The animation below shows the first 10 randomization replicates for both candidate models.


```{r}
#| label: fig-qqplot-animation
#| fig-cap: "Z-residual Q–Q plots for overall GOF (animation over 10 randomization replicates). Left: WBC model; right: log-WBC model. Each frame corresponds to one regenerated Z-residual set; the panel title reports the replicate-specific Z-SW p-value."
#| fig-align: center
#| code-summary: "GIF Generation Code (Folded)"
#| code-fold: true
#| fig-alt: "Animated QQ plots of Z-residuals comparing WBC and log-WBC Cox frailty models."

gif_qq_name <- "qqplot_anim.gif"
gif_qq_path <- file.path(extdata_path, gif_qq_name)

if (is_dev && (force_rerun || !file.exists(gif_qq_path))) {
  gifski::save_gif(
    expr = {
      for (i in 1:10) {
        par(mfrow = c(1, 2), mar = c(4, 4, 2, 2))
        qqnorm.zresid(Zresid.LeukSurv.wbc, irep = i)
        qqnorm.zresid(Zresid.LeukSurv.logwbc, irep = i)
      }
    },
    gif_file = gif_qq_path, width = 800, height = 400, res = 72, delay = 0.8
  )
}

local_qq <- get_local_asset(gif_qq_name, extdata_path, local_assets_dir)
if (!is.null(local_qq)) {
  knitr::include_graphics(local_qq)
} else {
  par(mfrow = c(1, 2))
  qqnorm.zresid(Zresid.LeukSurv.wbc, irep = 1)
  qqnorm.zresid(Zresid.LeukSurv.logwbc, irep = 1)
}
```

Across frames, both models show QQ patterns close to the diagonal with no obvious tail inflation. In this LeukSurv example, the overall GOF check alone does not distinguish the two competing functional forms (raw WBC vs log-WBC); the more targeted covariate-specific diagnostics in Sections 4.5–4.6 are needed.


### Assessing Homogeneity of Grouped Z-Residuals

A key Z-residual diagnostic for model adequacy is **homogeneity**: after sorting observations by the linear predictor (LP) and splitting into \(k\) groups, the grouped Z-residuals should have (approximately) the same mean and variance across LP groups if the model is adequate. Graphically, we expect:
- the LOWESS smooth in the Z-residual scatterplot vs LP to stay close to the horizontal line at 0, and
- the grouped boxplots to show similar centers and spreads.

We report two replicate-specific p-values to summarize the LP-group homogeneity checks. **Z-AOV-LP** is the ANOVA F-test p-value for equality of the *group means*, and **Z-BL-LP** is the Bartlett’s test p-value for equality of the *group variances*. In the LeukSurv example, across randomization replicates both the WBC and log-WBC models typically produce relatively large Z-AOV-LP and Z-BL-LP p-values, consistent with the visual impression that the LOWESS smooth stays near 0 and the grouped boxplots have comparable centers and spreads. Overall, LP-based grouping does not provide strong evidence of lack of fit for either model in this dataset.



```{r}
#| label: fig-zresid-scatter-box
#| fig-cap: "Homogeneity check by grouping Z-residuals over the linear predictor (LP) (animation over 10 randomization replicates). Top row: scatterplots with LOWESS smooth; bottom row: grouped boxplots. Each frame reports replicate-specific Z-AOV-LP (ANOVA for equal means) and Z-BL-LP (Bartlett’s test for equal variances) p-values."
#| code-summary: "GIF Generation Code (Folded)"
#| code-fold: true
#| fig-alt: "Animated scatterplots and boxplots of Z-residuals versus linear predictor for WBC and log-WBC models."

gif_lp_name <- "lp_anim.gif"
gif_lp_path <- file.path(extdata_path, gif_lp_name)

if (is_dev && (force_rerun || !file.exists(gif_lp_path))) {
  gifski::save_gif(
    expr = {
      for (i in 1:10) {
        par(mfrow = c(2, 2), mar = c(4, 4, 1.5, 2))
        plot(Zresid.LeukSurv.wbc, x_axis_var="lp", main.title = "Scatter: WBC Model", irep=i,add_lowess = TRUE)
        plot(Zresid.LeukSurv.logwbc, x_axis_var="lp", main.title = "Scatter: log-WBC Model", irep=i,add_lowess = TRUE)
        boxplot(Zresid.LeukSurv.wbc, x_axis_var = "lp", main.title = "Boxplot: WBC Model", irep=i)
        boxplot(Zresid.LeukSurv.logwbc, x_axis_var = "lp", main.title = "Boxplot: log-WBC Model", irep=i)
      }
    },
    gif_file = gif_lp_path, width = 900, height = 900, res = 96, delay = 1
  )
}

local_lp <- get_local_asset(gif_lp_name, extdata_path, local_assets_dir)
if (!is.null(local_lp)) knitr::include_graphics(local_lp)
```

In this dataset, the LP-based homogeneity diagnostic does not clearly separate the two candidate models. We therefore proceed to covariate-specific diagnostics to assess whether the functional form of the WBC effect is misspecified.

### Identifying Misspecification of Functional Form


To diagnose **functional-form misspecification** for a specific covariate, we inspect Z-residuals against that covariate (or its transformed version) and test whether grouped residual means differ across covariate intervals. If the functional form is adequate, the LOWESS curve should remain near 0 and grouped boxplots should have similar centers.

Here we directly compare the WBC model plotted against **wbc**, and the log-WBC (lwbc) model plotted against **logwbc**.

In the LeukSurv data, when diagnosing the functional form against log(wbc), the log-WBC model typically shows a pronounced systematic (non-linear) pattern in the LOWESS smooth and clear differences in grouped means, leading to consistently small **Z-AOV-log(wbc)** p-values across randomization replicates (e.g., < 0.01 in the real-data analysis). By contrast, the WBC model does not show comparably strong evidence of misspecification in this view: its replicated Z-AOV p-values are generally much larger, in line with a flatter smooth around zero and more homogeneous grouped boxplots. This contrast remains stable across many randomization replicates (see Sections 5.2–5.3).



```{r}
#| label: fig-zresid-scatter-box-wbc
#| fig-cap: "Covariate-specific functional-form diagnostic (animation over 10 randomization replicates). Left column (WBC model): Z-residuals vs wbc; right column (log-WBC model): Z-residuals vs logwbc. Top row: scatterplots with LOWESS smooth; bottom row: grouped boxplots with replicate-specific Z-AOV (covariate) p-values reported in the titles."
#| code-summary: "GIF Generation Code (Folded)"
#| code-fold: true
#| fig-alt: "Animated scatterplots and boxplots of Z-residuals versus WBC and log(WBC) for competing models."

gif_wbc_name <- "wbc_anim.gif"
gif_wbc_path <- file.path(extdata_path, gif_wbc_name)

if (is_dev && (force_rerun || !file.exists(gif_wbc_path))) {
  gifski::save_gif(
    expr = {
      for (i in 1:10) {
        par(mfrow = c(2, 2), mar = c(4, 4, 1.5, 2))
        plot(Zresid.LeukSurv.wbc, x_axis_var = "wbc", main.title = "Scatter: WBC Model",
        irep=i,add_lowess = TRUE)
        plot(Zresid.LeukSurv.logwbc, x_axis_var = "log(wbc)", main.title = "Scatter: log-WBC Model",          irep=i,add_lowess = TRUE)
        boxplot(Zresid.LeukSurv.wbc, x_axis_var = "wbc", main.title = "Boxplot: WBC Model", irep=i)
        boxplot(Zresid.LeukSurv.logwbc, x_axis_var = "log(wbc)", main.title = "Boxplot: log-WBC Model", irep=i)
      }
    },
    gif_file = gif_wbc_path, width = 900, height = 900, res = 96, delay = 1
  )
}

local_wbc <- get_local_asset(gif_wbc_name, extdata_path, local_assets_dir)
if (!is.null(local_wbc)) knitr::include_graphics(local_wbc)
```

## Statistical Test Summaries

### Quantitative Evaluation of Homogeneity

To quantify the homogeneity of grouped Z-residuals, we test whether the grouped residuals are “homogeneously distributed” by examining equality of group means using an ANOVA F-test (Z-AOV). In addition, we report Z-BL p-values from Bartlett’s test as a complementary check for equality of variances across the same groups.

Table 1 summarizes the first 10 randomization repetitions. In these repetitions, the LP-grouping diagnostics (Z-AOV-LP and Z-BL-LP) yield predominantly non-small p-values for both models, consistent with no strong evidence of non-homogeneity against the linear predictor. In contrast, the covariate-grouping test for the log-WBC specification (aov.lwbc) is consistently near zero, indicating strong group-mean differences across log(wbc) intervals.


```{r}
#| label: tbl-homo-first10
#| echo: false
#| tbl-cap: "Summary of residual homogeneity tests for the first 10 randomization repetitions. Z-AOV are ANOVA F-test p-values for equality of grouped Z-residual means; Z-BL are Bartlett’s test p-values for equality of variances. Suffix “.lp” indicates grouping by the linear predictor; columns without “.lp” indicate grouping by the covariate scale used in the diagnostic."

sw.wbc<-sw.test.zresid(Zresid.LeukSurv.wbc); sw.lwbc<-sw.test.zresid(Zresid.LeukSurv.logwbc)
sf.wbc<-sf.test.zresid(Zresid.LeukSurv.wbc); sf.lwbc<-sf.test.zresid(Zresid.LeukSurv.logwbc)

aov.wbc.lp<-aov.test.zresid(Zresid.LeukSurv.wbc, X="lp", k.anova=10)
aov.lwbc.lp<-aov.test.zresid(Zresid.LeukSurv.logwbc, X="lp", k.anova=10)

bl.wbc.lp<-bartlett.test.zresid(Zresid.LeukSurv.wbc, X="lp", k.bl=10)
bl.lwbc.lp<-bartlett.test.zresid(Zresid.LeukSurv.logwbc, X="lp", k.bl=10)

aov.wbc<-aov.test.zresid(Zresid.LeukSurv.wbc, X="wbc", k.anova=10)
aov.lwbc<-aov.test.zresid(Zresid.LeukSurv.logwbc, X="logwbc", k.anova=10)

bl.wbc<-bartlett.test.zresid(Zresid.LeukSurv.wbc, X="wbc", k.bl=10)
bl.lwbc<-bartlett.test.zresid(Zresid.LeukSurv.logwbc, X="logwbc", k.bl=10)

homogeneity_tests <- data.frame(aov.wbc.lp, aov.lwbc.lp, bl.wbc.lp, bl.lwbc.lp, aov.wbc, aov.lwbc, bl.wbc, bl.lwbc)

homogeneity_tests %>% head(10) %>% gt() %>% tab_header(title = "Summary of Residual Homogeneity Tests") %>% fmt_number(columns = everything(), decimals = 4)
```

### Summary of Minimum P-values ($p_{min}$)
To connect the above visuals to the replicated-test summary used in the paper, the following compact table reproduces the LeukSurv comparison (AIC, CZ-CSF p-value, and $p_{min}$ summaries over 1000 replicates). Notably, the log-WBC model fails the Z-AOV-log(wbc) diagnostic decisively, even though some overall-fit diagnostics (e.g., CZ-CSF and normality-based tests) do not flag this inadequacy.
```{r pmin-table, echo=FALSE}
#| label: tbl-leuksurv-summary
#| echo: false
#| tbl-cap: "LeukSurv model comparison and replicated-test summaries (AIC, CZ-CSF p-value, and $p_{min}$ across 1000 randomization replicates)."

leuksurv_summary <- tibble::tribble(
~Model, ~AIC, ~`CZ-CSF (p-value)`, ~`Z-SW (p_min)`, ~`Z-SF (p_min)`, ~`Z-AOV-LP (p_min)`, ~`Z-AOV-log(wbc) (p_min)`,
"WBC model",     3111.669, 0.255, 0.495, 0.693, 0.703, 0.074,
"log-WBC model", 3132.105, 0.305, 0.579, 0.781, 0.978, NA_real_
) %>%
dplyr::mutate(`Z-AOV-log(wbc) (p_min)` = dplyr::if_else(Model == "log-WBC model", NA_real_, `Z-AOV-log(wbc) (p_min)`))

leuksurv_summary %>%
gt::gt() %>%
gt::fmt_number(columns = c(AIC, `CZ-CSF (p-value)`, `Z-SW (p_min)`, `Z-SF (p_min)`, `Z-AOV-LP (p_min)`, `Z-AOV-log(wbc) (p_min)`),
decimals = 3) %>%
gt::text_transform(
locations = gt::cells_body(columns = `Z-AOV-log(wbc) (p_min)`, rows = Model == "log-WBC model"),
fn = function(x) rep("<1e-5", length(x))
) %>%
gt::tab_source_note("$p_{min}$ is a conservative upper-bound summary for replicated p-values; a practical failure cutoff can be much larger than 0.05 (e.g., 0.25).")

```
### Histograms of Replicated P-values

These histograms reveal that the Z-SW, Z-SF, and Z-AOV-LP tests for both models have a substantial proportion of p-values greater than 0.05, leading to large pmin values. In contrast, the replicated Z-AOV-log(wbc) p-values for the lwbc model are nearly all smaller than 0.001, providing strong evidence that the log transformation of wbc is inappropriate for modelling the survival time.

```{r}
#| label: fig-zresid-hist-pvalues
#| fig-height: 9
#| fig-width: 9
#| fig-align: center
#| fig-cap: "Histograms of 1000 replicated Z-residual test p-values for the wbc model (left panels) and the lwbc model (right panels). The vertical red lines indicate pmin for the 1000 replicated p-values. Note that the upper limit of the x-axis for Z-AOV-log(wbc) p-values for the lwbc model is 0.005, and 1 for others."


pmin.sw.wbc<-pvalue.min(pv=sw.wbc); pmin.sw.lwbc<-pvalue.min(pv=sw.lwbc)
pmin.sf.wbc<-pvalue.min(pv=sf.wbc); pmin.sf.lwbc<-pvalue.min(pv=sf.lwbc)
pmin.aov.lp.wbc<-pvalue.min(pv=aov.wbc.lp); pmin.aov.lp.lwbc<-pvalue.min(pv=aov.lwbc.lp)
pmin.aov.wbc<-pvalue.min(pv=aov.wbc); pmin.aov.lwbc<-pvalue.min(pv=aov.lwbc)

par(mfrow = c(4, 2), mar = c(4, 4, 2, 2))
hist(sw.wbc, main = "Z-SW: WBC Model", breaks = 20); abline(v = pmin.sw.wbc, col = "red")
hist(sw.lwbc, main = "Z-SW: log-WBC Model", breaks = 20); abline(v = pmin.sw.lwbc, col = "red")
hist(sf.wbc, main = "Z-SF: WBC Model", breaks = 20); abline(v = pmin.sf.wbc, col = "red")
hist(sf.lwbc, main = "Z-SF: log-WBC Model", breaks = 20); abline(v = pmin.sf.lwbc, col = "red")
hist(aov.wbc.lp, main = "Z-AOV (LP): WBC Model", breaks = 20); abline(v = pmin.aov.lp.wbc, col = "red")
hist(aov.lwbc.lp, main = "Z-AOV (LP): log-WBC Model", breaks = 20); abline(v = pmin.aov.lp.lwbc, col = "red")
hist(aov.wbc, main = "Z-AOV (WBC): WBC Model", breaks = 20); abline(v = pmin.aov.wbc, col = "red")
hist(aov.lwbc, main = "Z-AOV (log-WBC): log-WBC Model", breaks = 20); abline(v = pmin.aov.lwbc, col = "red")
```


## Other Residual Analysis

### Censored Z-residuals
We first report the **CZ-CSF** p-values, where “censored Z-residuals” are defined by
$$
r^{n}_{ij}(t_{ij}) = -\Phi^{-1}\!\left(\widehat{S}_{ij}(t_{ij})\right),
$$
and normality is assessed using the SF test for multiply censored data (EnvStats::gofTestCensored).  

As an overall goodness-of-fit check, CZ-CSF evaluates the residual distribution but does not assess specific model assumptions such as the functional form of covariates.

```{r}
#| label: tbl-czcsf
#| echo: false
#| tbl-cap: "CZ-CSF p-values (SF normality test for censored Z-residuals) for the wbc and lwbc models."

censored.Zresid.wbc <- surv_residuals(
  fit.object = fit_LeukSurv_wbc, data = LeukSurv,
  residual.type = "censored Z-residual"
)
censored.Zresid.lwbc <- surv_residuals(
  fit.object = fit_LeukSurv_logwbc, data = LeukSurv,
  residual.type = "censored Z-residual"
)

gof_wbc  <- gof.censored.zresidual(censored.Zresidual = censored.Zresid.wbc)
gof_lwbc <- gof.censored.zresidual(censored.Zresidual = censored.Zresid.lwbc)

extract_pvalue <- function(obj) {
  if (inherits(obj, "htest")) return(as.numeric(obj$p.value))
  if (is.list(obj)) {
    for (k in c("p.value", "p_value", "pvalue", "p")) {
      if (!is.null(obj[[k]])) return(as.numeric(obj[[k]][1]))
    }
  }
  if (is.numeric(obj) && length(obj) == 1) return(as.numeric(obj))
  txt <- paste(capture.output(print(obj)), collapse = "\n")
  m <- regexpr("p[- ]value\\s*[:=]\\s*([0-9\\.eE+-]+)", txt, perl = TRUE, ignore.case = TRUE)
  if (m[1] != -1) {
    s <- regmatches(txt, m)
    return(as.numeric(sub(".*p[- ]value\\s*[:=]\\s*", "", s, ignore.case = TRUE)))
  }
  NA_real_
}

czcsf_tbl <- tibble::tibble(
  Model = c("wbc model", "lwbc model"),
  `CZ-CSF p-value` = c(extract_pvalue(gof_wbc), extract_pvalue(gof_lwbc))
)

czcsf_tbl %>%
  gt::gt() %>%
  gt::fmt_number(columns = `CZ-CSF p-value`, decimals = 4)

```



### Cox-Snell Residuals
The Cox–Snell residual is defined as
$$
r^{CS}_{ij}(t_{ij}) = -\log\!\left(\widehat{S}_{ij}(t_{ij})\right).
$$

Overall GOF checks based on CS residuals are commonly used, but they may not provide sufficient information about specific model inadequacies (e.g., the functional form of a covariate).

```{r Cox-Snell Residuals}
#| label: fig-Cox-Snell_Residuals
#| fig-cap: "Cumulative hazard plot of Cox–Snell (CS) residuals for the wbc and lwbc models. With censoring, the Kaplan–Meier estimate of the survivor function of CS residuals is compared against the 45° straight line as an overall diagnostic."
#| fig-align: center
#| out-width: "100%"

ucs.wbc <- surv_residuals(fit.object = fit_LeukSurv_wbc, data= LeukSurv, residual.type = "Cox-Snell" )
ucs.lwbc <- surv_residuals(fit.object = fit_LeukSurv_logwbc, data= LeukSurv, residual.type = "Cox-Snell" )
par(mfrow = c(1, 2)); plot.cs.residual(ucs.wbc, main.title = "CS Residuals: WBC Model"); plot.cs.residual(ucs.lwbc, main.title = "CS Residuals: log-WBC Model")
```

### Martingale Residuals
To gain further insights, residual scatterplots are often augmented with LOWESS lines; however, the visual interpretation can still be subjective, and numerical tests are difficult to derive because these residuals lack a reference distribution under censoring.

```{r martingale-residuals}
#| label: fig-Martingale_Residuals
#| fig-cap: "Martingale residual plots against wbc and log(wbc) for the wbc and lwbc models, with a LOWESS smooth. Martingale residuals are commonly used for visually assessing covariate functional form, but they are asymmetric and lack a lower bound; moreover, deriving numerical tests is challenging due to censoring and the lack of a reference distribution."
#| fig-align: center
#| out-width: "100%"

martg.wbc <- surv_residuals(fit.object = fit_LeukSurv_wbc, data= LeukSurv, residual.type = "martingale")
martg.lwbc <- surv_residuals(fit.object = fit_LeukSurv_logwbc, data= LeukSurv, residual.type = "martingale")
par(mfrow = c(1, 2))
plot.martg.resid(martg.wbc, x_axis_var="wbc", main.title = "Martingale Residuals: WBC Model")
plot.martg.resid(martg.lwbc, x_axis_var="logwbc", main.title = "Martingale Residuals: log-WBC Model")
```

### Deviance Residuals
Although deviance residuals reduce skewness relative to martingale residuals and can improve visual assessment, objective numerical measures remain hard to construct under censoring because these residuals do not admit a convenient reference distribution. 

```{r deviance-residuals}
#| label: fig-Deviance_Residuals
#| fig-cap: "Deviance residual plots against wbc and log(wbc) for the wbc and lwbc models, with a LOWESS smooth. Deviance residuals are a normalized transformation of martingale residuals that aim to achieve symmetry; when the fitted model is appropriate, they have mean zero and are approximately symmetrically distributed around zero."
#| fig-align: center
#| out-width: "100%"
dev.wbc <- surv_residuals(fit.object = fit_LeukSurv_wbc, data= LeukSurv, residual.type = "deviance")
dev.lwbc <- surv_residuals(fit.object = fit_LeukSurv_logwbc, data= LeukSurv, residual.type = "deviance")
par(mfrow = c(1, 2))
plot.dev.resid(dev.wbc, x_axis_var="wbc", main.title = "Deviance Residuals: WBC Model")
plot.dev.resid(dev.lwbc, x_axis_var="logwbc", main.title = "Deviance Residuals: log-WBC Model")
```

## References
Wu, T., Li, L., & Feng, C. (2024). Z-residual diagnostic tool for assessing covariate functional form in shared frailty models. *Journal of Applied Statistics*, 52(1), 28–58. <https://doi.org/10.1080/02664763.2024.2355551>
