---
title: "Z-residual Diagnostics for `brms` Logistic Regression"
author: "Tingxuan Wu and Longhai Li"
execute: 
  cache: false
  message: false
format: 
  html:
    number-sections: true
    code-fold: true
vignette: >
  %\VignetteIndexEntry{Demo of Zresidual API Methods for brms Logistic Regression}
  %\VignetteEncoding{UTF-8}
  %\VignetteEngine{quarto::html}
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(
  fig.width = 7,
  fig.height = 5,
  fig.align = "center"
)

# Define the global toggles to force regeneration of computations
force_refit <- FALSE   # Controls model caching
force_recalcz <- FALSE # Controls Z-residual calculations for power analysis

suppressWarnings(unloadNamespace("Zresidual"))
library(Zresidual)
library(brms)
library(knitr)
library(kableExtra)

#Check if cmdstanr is available
# Check for the R package and the C++ backend
if (!requireNamespace("cmdstanr", quietly = TRUE) || is.null(cmdstanr::cmdstan_path())) {
  message("The 'cmdstanr' package and/or CmdStan toolchain are required.")
  message("1. Install R package: install.packages('cmdstanr', repos = 'https://mc-stan.org/r-packages/')")
  message("2. Install C++ backend: cmdstanr::install_cmdstan()")
  stop("Missing cmdstanr dependencies.", call. = FALSE)
}

model_dir <- "resources/brms_bernoulli_models"
dir.create(model_dir, recursive = TRUE)

```

## Preparation: Data and Model Caching

This vignette demonstrates the three distinct ways you can use `Zresidual` to evaluate Bayesian models fit via `brms`:

1. Passing a **Custom User-Defined Function**.
2. Triggering the **Simulation-Predictive Function**.
3. Utilizing the **Built-in Analytic Predictive Function**.

To illustrate these methods effectively, we establish a consistent testing environment. We assume your environment already contains a pre-loaded list of 100 datasets (`sim_data`), where the true log-odds follow a non-linear sine wave. We fit two models to each dataset: a misspecified linear model and a correctly specified non-linear model. 

Because MCMC sampling is computationally expensive, we execute and cache these 200 model fits to disk. In the subsequent sections, we will instantly load these identical models to demonstrate how the three `Zresidual` calling methods yield identical, reliable diagnostics.

```{r prep-data-and-models}
#| code-fold: true
#| message: false
#| warning: false

if (!dir.exists(model_dir)) {
  dir.create(model_dir, recursive = TRUE)
}

n_sim <- 100
n_sample <- 200 
sim_data <- vector("list", n_sim)

# Loop through iterations, checking for existing files before simulating/fitting
for(i in 1:n_sim) {
  
  data_file <- paste0(model_dir, "/data_sim_", i, ".rds")
  wrong_model_file <- paste0(model_dir, "/wrong_sim_", i, ".rds")
  correct_model_file <- paste0(model_dir, "/correct_sim_", i, ".rds")
  
  # --- 1. Check and Handle Data ---
  if (!force_refit && file.exists(data_file)) {
    sim_data[[i]] <- readRDS(data_file)
  } else {
    # First-time running: Simulate the Bernoulli data
    x <- runif(n_sample, 0, 10)
    eta <- 2.5 * sin(x)
    y <- rbinom(n_sample, 1, plogis(eta))
    
    dat_i <- data.frame(y = y, x = x)
    sim_data[[i]] <- dat_i
    saveRDS(dat_i, file = data_file)
  }
  
  # --- 2. Check and Fit Wrong Model ---
  if (force_refit || !file.exists(wrong_model_file)) {
    brm(
      y ~ x, 
      family = bernoulli(), 
      data = sim_data[[i]],
      backend = "cmdstanr",
      normalize = FALSE,
      file = paste0(model_dir, "/wrong_sim_", i), 
      refresh = 0,
      cores = 4
    )
  }
  
  # --- 3. Check and Fit Correct Model ---
  if (force_refit || !file.exists(correct_model_file)) {
    brm(
      y ~ sin(x), 
      family = bernoulli(), 
      data = sim_data[[i]],
      backend = "cmdstanr",
      normalize = FALSE,
      file = paste0(model_dir, "/correct_sim_", i), 
      refresh = 0,
      cores = 4
    )
  }
}
```

## `Zresidual` Using a Custom User-Defined `log_pointpred` Function

`Zresidual` is built with a highly extensible architecture. If your specific model family or backend is not natively supported yet, or if you simply want explicit control over the math, you can bypass the internal defaults entirely by passing a custom function to the `log_pointpred` argument.

Below, we define a custom analytical function that leverages `brms::posterior_epred()` to manually extract the expected probabilities ($p$) and push them through R's native distribution functions.

### Defining a Custom Predictive Function

```{r custom-log-pointpred}
#| eval: true 
#| code-fold: show

log_pointpred_brms_bern <- function(fit, data = NULL, ...) {
  
  # 1. Resolve data
  data_in <- if (is.null(data)) fit$data else data
  
  # 2. Extract observed response safely
  model.data <- model.frame(fit$formula, data = data_in)
  model.var  <- names(model.data)
  
  response <- fit$formula$resp
  if (is.null(response) || !response %in% model.var) {
    response <- all.vars(fit$formula$formula)[1]
  }
  
  y_obs <- as.numeric(model.data[[response]])
  n <- length(y_obs)
  
  # 3. Extract Posterior Expected Values (probabilities)
  p <- brms::posterior_epred(fit, dpar = "mu", newdata = data_in, ...)
  mc_used <- nrow(p)
  
  # 4. Calculate Log-PMF and Log-Survival Functions
  lpmf <- matrix(NA_real_, mc_used, n)
  lsf  <- matrix(NA_real_, mc_used, n)
  
  for (i in seq_len(n)) {
    lpmf[, i] <- dbinom(y_obs[i], size = 1, prob = p[, i], log = TRUE)
    lsf[, i]  <- pbinom(y_obs[i], size = 1, prob = p[, i],
                           lower.tail = FALSE, log.p = TRUE)
  }
  
  list(
    log_like = lpmf,
    log_surv  = lsf,
    is_discrete = rep(1L, n)
  )
}
```

### Single Dataset Evaluation

To evaluate our model, we explicitly map the custom function to the `log_pointpred` argument. `Zresidual` handles the rest of the randomization, prediction extraction, and covariate binding automatically.

```{r}
set.seed(123) ## for reproducibility
```

```{r user-single-model-eval}
#| code-fold: show
#| results: hide
#| message: false

# Load the first cached dataset and model pair
dat <- readRDS(paste0(model_dir, "/data_sim_1.rds"))

fit_wrong <- brm(y ~ x, family = bernoulli(), data = dat, backend = "cmdstanr", file = paste0(model_dir, "/wrong_sim_1"))
fit_correct <- brm(y ~ sin(x), family = bernoulli(), data = dat, backend = "cmdstanr", file = paste0(model_dir, "/correct_sim_1"))

# Call Zresidual explicitly using the custom function
z_wrong   <- Zresidual(fit_wrong, data = dat, mcmc_summarize = "post", randomized = TRUE, nrep = 10, log_pointpred = log_pointpred_brms_bern)
z_correct <- Zresidual(fit_correct, data = dat, mcmc_summarize = "post", randomized = TRUE, nrep = 10, log_pointpred = log_pointpred_brms_bern)
```

### Diagnostics with Z-residuals

```{r user-diagnosis-plots}
#| code-fold: true
#| fig-width: 12
#| fig-height: 8

i <- 1 
par(mfrow = c(2, 3), mar = c(4, 4, 3, 1))

qqnorm(z_wrong, irep = i, main = "Wrong: Z-Resid Q-Q")
plot(z_wrong, x_axis_var = "x", category = dat$y, irep = i, main = "Wrong: Z-Resid vs X")
boxplot(z_wrong, x_axis_var = "lp", irep = i, main = "Wrong: Z-Resid vs LP")

qqnorm(z_correct, irep = i, main = "Correct: Z-Resid Q-Q")
plot(z_correct, x_axis_var = "x", category = dat$y, irep = i, main = "Correct: Z-Resid vs X")
boxplot(z_correct, x_axis_var = "lp", irep = i, main = "Correct: Z-Resid vs LP")
```

**Replicated p-values of the same fitted models**
```{r user-replicated-p-values-spanned}
#| code-fold: true

res_table <- data.frame(
  "Replicate" = paste("Rep", 1:ncol(z_wrong)),
  "SW_W"      = as.numeric(sw.test.zresid(z_wrong)),
  "AOV_W"     = as.numeric(aov.test.zresid(z_wrong, X = "lp")),
  "BL_W"      = as.numeric(bartlett.test.zresid(z_wrong, X = "lp")),
  "SW_C"      = as.numeric(sw.test.zresid(z_correct)),
  "AOV_C"     = as.numeric(aov.test.zresid(z_correct, X = "lp")),
  "BL_C"      = as.numeric(bartlett.test.zresid(z_correct, X = "lp"))
)

res_table %>%
  kable(
    digits = 4, 
    col.names = c("Replicate", "SW", "AOV", "Bartlett", "SW", "AOV", "Bartlett"),
    caption = "Diagnostic Test p-values Across Randomized Replicates",
    align = "lcccccc"
  ) %>%
  kable_styling(
    bootstrap_options = c("striped", "condensed"), full_width = FALSE, position = "center"
  ) %>%
  add_header_above(c(" " = 1, "Wrong Model (Linear)" = 3, "Correct Model (Non-linear)" = 3))
```

### Checking the Sampling Distribution of ANOVA Nonlinearity Tests p-values

By batching this custom `Zresidual` call across our 100 simulated datasets, we construct a sampling distribution. The uniform $p$-values for the correctly specified model confirm that providing a user-defined function seamlessly integrates with the `Zresidual` statistical pipeline.

```{r user-replication-verify-anova}
#| code-fold: true
#| results: hide
#| message: false
#| warning: false

user_res_file <- paste0(model_dir, "/user_pvalues.rds")

if (!force_recalcz && file.exists(user_res_file)) {
  res <- readRDS(user_res_file)
  p_c <- res$p_c
  p_w <- res$p_w
} else {
  p_c <- numeric(n_sim)
  p_w <- numeric(n_sim)
  
  for(i in 1:n_sim) {
    dat_sim <- readRDS(paste0(model_dir, "/data_sim_", i, ".rds"))
    fit_w_sim <- brm(y ~ x, family = bernoulli(), data = dat_sim, backend = "cmdstanr", file = paste0(model_dir, "/wrong_sim_", i), refresh = 0)
    z_w_sim <- Zresidual(fit_w_sim, data = dat_sim, mcmc_summarize = "post", randomized = TRUE, nrep = 1, log_pointpred = log_pointpred_brms_bern)
    p_w[i] <- as.numeric(aov.test.zresid(z_w_sim, X = "lp"))
    
    fit_c_sim <- brm(y ~ sin(x), family = bernoulli(), data = dat_sim, backend = "cmdstanr", file = paste0(model_dir, "/correct_sim_", i), refresh = 0)
    z_c_sim <- Zresidual(fit_c_sim, data = dat_sim, mcmc_summarize = "post", randomized = TRUE, nrep = 1, log_pointpred = log_pointpred_brms_bern)
    p_c[i] <- as.numeric(aov.test.zresid(z_c_sim, X = "lp"))
  }
  saveRDS(list(p_c = p_c, p_w = p_w), user_res_file)
}
```

```{r user-replication-hists}
#| fig-width: 10
#| fig-height: 5
#| code-fold: true

par(mfrow = c(1, 2))
hist(p_c, main = "Correct: Model P-values", xlab = "p-value", col = "lightblue", breaks = 10)
abline(h = n_sim / 10, col = "red", lty = 2)

hist(p_w, main = "Wrong: Small p-values", xlab = "p-value", col = "salmon", breaks = 10)
```


## `Zresidual` Using the Simulation-Predictive Method

When an exact analytic mathematical formula is either unknown or computationally prohibitive for a specific model family, `Zresidual` provides a robust simulation mechanism: the **Simulation-Predictive** method. 

By passing `pred_method = "simulation"`, `Zresidual` automatically invokes the model's native `simulate` or `posterior_predict` functions to draw empirical bounds, bypassing the need for explicit PDF/CDF math entirely.

### Single Dataset Evaluation

Notice how we drop the `log_pointpred` argument here and explicitly request the `"simulation"` method. `Zresidual` automatically handles the translation from empirical draws to normalized residuals.

```{r}
set.seed(123) ## for reproducibility
```

```{r sim-single-model-eval}
#| code-fold: show
#| results: hide
#| message: false

dat <- readRDS(paste0(model_dir, "/data_sim_1.rds"))
fit_wrong <- brm(y ~ x, family = bernoulli(), data = dat, backend = "cmdstanr", file = paste0(model_dir, "/wrong_sim_1"))
fit_correct <- brm(y ~ sin(x), family = bernoulli(), data = dat, backend = "cmdstanr", file = paste0(model_dir, "/correct_sim_1"))

# Trigger the Simulation Fallback method
z_wrong   <- Zresidual(fit_wrong, data = dat, pred_method = "simulation", mcmc_summarize = "iscv", randomized = TRUE, nrep = 10)
z_correct <- Zresidual(fit_correct, data = dat, pred_method = "simulation", mcmc_summarize = "iscv", randomized = TRUE, nrep = 10)
```

### Diagnostics with Z-residuals

```{r sim-diagnosis-plots}
#| code-fold: true
#| fig-width: 12
#| fig-height: 8

i <- 1 
par(mfrow = c(2, 3), mar = c(4, 4, 3, 1))

qqnorm(z_wrong, irep = i, main = "Wrong: Z-Resid Q-Q")
plot(z_wrong, x_axis_var = "x", category = dat$y, irep = i, main = "Wrong: Z-Resid vs X")
boxplot(z_wrong, x_axis_var = "lp", irep = i, main = "Wrong: Z-Resid vs LP")

qqnorm(z_correct, irep = i, main = "Correct: Z-Resid Q-Q")
plot(z_correct, x_axis_var = "x", category = dat$y, irep = i, main = "Correct: Z-Resid vs X")
boxplot(z_correct, x_axis_var = "lp", irep = i, main = "Correct: Z-Resid vs LP")
```

**Replicated p-values of the same fitted models**
```{r sim-replicated-p-values-spanned}
#| code-fold: true

res_table <- data.frame(
  "Replicate" = paste("Rep", 1:ncol(z_wrong)),
  "SW_W"      = as.numeric(sw.test.zresid(z_wrong)),
  "AOV_W"     = as.numeric(aov.test.zresid(z_wrong, X = "lp")),
  "BL_W"      = as.numeric(bartlett.test.zresid(z_wrong, X = "lp")),
  "SW_C"      = as.numeric(sw.test.zresid(z_correct)),
  "AOV_C"     = as.numeric(aov.test.zresid(z_correct, X = "lp")),
  "BL_C"      = as.numeric(bartlett.test.zresid(z_correct, X = "lp"))
)

res_table %>%
  kable(
    digits = 4, 
    col.names = c("Replicate", "SW", "AOV", "Bartlett", "SW", "AOV", "Bartlett"),
    caption = "Diagnostic Test p-values Across Randomized Replicates",
    align = "lcccccc"
  ) %>%
  kable_styling(
    bootstrap_options = c("striped", "condensed"), full_width = FALSE, position = "center"
  ) %>%
  add_header_above(c(" " = 1, "Wrong Model (Linear)" = 3, "Correct Model (Non-linear)" = 3))
```

### Checking the Sampling Distribution of ANOVA Nonlinearity Tests p-values

Batching this approach confirms that the simulation fallback is statistically valid. It generates the identical expected behaviors (uniform distributions for correct models, skewed distributions for structural misspecifications) without requiring a single line of custom analytical code.

```{r sim-sampling-distribution}
#| code-fold: true
#| results: hide
#| message: false
#| warning: false

sim_res_file <- paste0(model_dir, "/sim_pvalues.rds")

if (!force_recalcz && file.exists(sim_res_file)) {
  res <- readRDS(sim_res_file)
  p_c <- res$p_c
  p_w <- res$p_w
} else {
  p_c <- numeric(n_sim)
  p_w <- numeric(n_sim)

  for(i in 1:n_sim) {
    dat_sim <- readRDS(paste0(model_dir, "/data_sim_", i, ".rds"))
    
    fit_w_sim <- brm(y ~ x, family = bernoulli(), data = dat_sim, backend = "cmdstanr", file = paste0(model_dir, "/wrong_sim_", i))
    z_w_sim <- Zresidual(fit_w_sim, data = dat_sim, pred_method = "simulation", mcmc_summarize = "iscv", randomized = TRUE, nrep = 1)
    p_w[i] <- as.numeric(aov.test.zresid(z_w_sim, X = "lp"))
    
    fit_c_sim <- brm(y ~ sin(x), family = bernoulli(), data = dat_sim, backend = "cmdstanr", file = paste0(model_dir, "/correct_sim_", i))
    z_c_sim <- Zresidual(fit_c_sim, data = dat_sim, pred_method = "simulation", mcmc_summarize = "iscv", randomized = TRUE, nrep = 1)
    p_c[i] <- as.numeric(aov.test.zresid(z_c_sim, X = "lp"))
  }
  saveRDS(list(p_c = p_c, p_w = p_w), sim_res_file)
}
```

```{r sim-hist-plots}
#| code-fold: true
#| fig-width: 10
#| fig-height: 5

par(mfrow = c(1, 2))
hist(p_c, main = "Correct: Model P-values", xlab = "p-value", col = "lightblue", breaks = 10)
abline(h = n_sim / 10, col = "red", lty = 2, lwd = 2)

hist(p_w, main = "Wrong: Misspecification P-values", xlab = "p-value", col = "salmon", breaks = 10)
```


## `Zresidual` Using the Built-In Analytical Routing

For standard model families evaluated on fully supported backends (like `brms` and standard `glm`), `Zresidual` aims to require zero configuration. 

By simply omitting the predictive arguments entirely, `Zresidual` automatically queries the model object's class and family string. It then silently routes the computation to highly optimized, natively written routines embedded directly within the package itself.

### Single Dataset Evaluation

The simplest—and generally fastest—way to use `Zresidual` is to provide only the fitted model object and the data.

```{r}
set.seed(123) ## for reproducibility
```

```{r ana-single-model-eval}
#| code-fold: show
#| results: hide
#| message: false

dat <- readRDS(paste0(model_dir, "/data_sim_1.rds"))
fit_wrong <- brm(y ~ x, family = bernoulli(), data = dat, backend = "cmdstanr", file = paste0(model_dir, "/wrong_sim_1"))
fit_correct <- brm(y ~ sin(x), family = bernoulli(), data = dat, backend = "cmdstanr", file = paste0(model_dir, "/correct_sim_1"))

# Calculate Z-residuals using default native routing
z_wrong   <- Zresidual(fit_wrong, data = dat, mcmc_summarize = "post", randomized = TRUE, nrep = 10)
z_correct <- Zresidual(fit_correct, data = dat, mcmc_summarize = "post", randomized = TRUE, nrep = 10)
```

### Diagnostics with Z-residuals

```{r ana-diagnosis-plots}
#| code-fold: true
#| fig-width: 12
#| fig-height: 8

i <- 1 
par(mfrow = c(2, 3), mar = c(4, 4, 3, 1))

qqnorm(z_wrong, irep = i, main = "Wrong: Z-Resid Q-Q")
plot(z_wrong, x_axis_var = "x", category = dat$y, irep = i, main = "Wrong: Z-Resid vs X")
boxplot(z_wrong, x_axis_var = "lp", irep = i, main = "Wrong: Z-Resid vs LP")

qqnorm(z_correct, irep = i, main = "Correct: Z-Resid Q-Q")
plot(z_correct, x_axis_var = "x", category = dat$y, irep = i, main = "Correct: Z-Resid vs X")
boxplot(z_correct, x_axis_var = "lp", irep = i, main = "Correct: Z-Resid vs LP")
```

**Replicated p-values of the same fitted models**
```{r ana-replicated-p-values-spanned}
#| code-fold: true

res_table <- data.frame(
  "Replicate" = paste("Rep", 1:ncol(z_wrong)),
  "SW_W"      = as.numeric(sw.test.zresid(z_wrong)),
  "AOV_W"     = as.numeric(aov.test.zresid(z_wrong, X = "lp")),
  "BL_W"      = as.numeric(bartlett.test.zresid(z_wrong, X = "lp")),
  "SW_C"      = as.numeric(sw.test.zresid(z_correct)),
  "AOV_C"     = as.numeric(aov.test.zresid(z_correct, X = "lp")),
  "BL_C"      = as.numeric(bartlett.test.zresid(z_correct, X = "lp"))
)

res_table %>%
  kable(
    digits = 4, 
    col.names = c("Replicate", "SW", "AOV", "Bartlett", "SW", "AOV", "Bartlett"),
    caption = "Diagnostic Test p-values Across Randomized Replicates",
    align = "lcccccc"
  ) %>%
  kable_styling(
    bootstrap_options = c("striped", "condensed"), full_width = FALSE, position = "center"
  ) %>%
  add_header_above(c(" " = 1, "Wrong Model (Linear)" = 3, "Correct Model (Non-linear)" = 3))
```

### Checking the Sampling Distribution of ANOVA Nonlinearity Tests p-values

Our final replication confirms that the native built-in routing produces the exact same robust diagnostic sampling distributions as our custom extraction logic, achieving uniform p-values for properly specified models.

```{r ana-replication-verify-anova}
#| code-fold: true
#| results: hide
#| message: false
#| warning: false

ana_res_file <- paste0(model_dir, "/ana_pvalues.rds")

if (!force_recalcz && file.exists(ana_res_file)) {
  res <- readRDS(ana_res_file)
  p_c <- res$p_c
  p_w <- res$p_w
} else {
  p_c <- numeric(n_sim)
  p_w <- numeric(n_sim)
  
  for(i in 1:n_sim) {
    dat_sim <- readRDS(paste0(model_dir, "/data_sim_", i, ".rds"))
    
    fit_w_sim <- brm(y ~ x, family = bernoulli(), data = dat_sim, backend = "cmdstanr", file = paste0(model_dir, "/wrong_sim_", i), refresh = 0)
    z_w_sim <- Zresidual(fit_w_sim, data = dat_sim, mcmc_summarize = "post", randomized = TRUE, nrep = 1)
    p_w[i] <- as.numeric(aov.test.zresid(z_w_sim, X = "lp"))
    
    fit_c_sim <- brm(y ~ sin(x), family = bernoulli(), data = dat_sim, backend = "cmdstanr", file = paste0(model_dir, "/correct_sim_", i), refresh = 0)
    z_c_sim <- Zresidual(fit_c_sim, data = dat_sim, mcmc_summarize = "post", randomized = TRUE, nrep = 1)
    p_c[i] <- as.numeric(aov.test.zresid(z_c_sim, X = "lp"))
  }
  
  saveRDS(list(p_c = p_c, p_w = p_w), ana_res_file)
}
```

```{r ana-replication-hists}
#| fig-width: 10
#| fig-height: 5
#| code-fold: true

par(mfrow = c(1, 2))
hist(p_c, main = "Correct: Model P-values", xlab = "p-value", col = "lightblue", breaks = 10)
abline(h = n_sim / 10, col = "red", lty = 2)

hist(p_w, main = "Wrong: Small p-values", xlab = "p-value", col = "salmon", breaks = 10)
```
