---
title: "Z-residual Diagnostics for `glm` Logistic Regression"
author: "Tingxuan Wu and Longhai Li"
format: 
  html:
    number-sections: false
    code-fold: true
vignette: >
  %\VignetteIndexEntry{Demo of Z-residual diagnostics for logistic regression via glm}
  %\VignetteEncoding{UTF-8}
  %\VignetteEngine{quarto::html}
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(
  fig.width = 7,
  fig.height = 5,
  fig.align = "center"
)
set.seed(2026)

# Set TRUE if you want to force regeneration of heavy computations
force_rerun <- FALSE


# Prefix string for all saved files
file_prefix <- "resources/glm_binomial/"
dir.create(file_prefix, recursive = TRUE)


```

```{r load-zresidual}
#| code-fold: show
#| message: false
#| warning: false

suppressWarnings(unloadNamespace("Zresidual"))
library(Zresidual)
```

## 1. Simulating a Non-linear Dataset

Because `Zresidual` provides native support for models fitted via the base R `glm()` function, we do not need to construct a custom predictive function. We can jump straight into simulating our dataset.

We simulate a dataset where the true log-odds follow a sine wave. With $x \in [0, 10]$, the average log-odds are near zero, naturally keeping prevalence $\approx 0.5$.

```{r}
#| label: simulate-data
#| code-fold: show

n <- 200

x <- runif(n, 0, 10)
# True non-linear relationship (sine wave)
eta_true <- 2.5 * sin(x) 
y <- rbinom(n, 1, plogis(eta_true))

dat <- data.frame(y = y, x = x)

cat("Prevalence (proportion of 1s):", mean(dat$y), "\n")
```

## 2. Fitting Models and Calculating Z-residuals

We compare a **Wrong Model** (assuming a linear relationship) against a **Correct Model** (specifying the sine term).

```{r}
#| label: fit-and-calc

# 1. Wrong Model (Linear)
fit_wrong <- glm(y ~ x, family = binomial(), data = dat)

# 2. Correct Model (Non-linear)
fit_correct <- glm(y ~ sin(x), family = binomial(), data = dat)

# Calculate Z-residuals using the built-in predictive backend (10 replicates each). 
# Covariates (zcov) are computed internally.
z_wrong   <- Zresidual(fit_wrong, data = dat, randomized = TRUE, nrep = 10)
z_correct <- Zresidual(fit_correct, data = dat, randomized = TRUE, nrep = 10)
```

## 3. Diagnostics with Z-residuals

The plots below display the diagnostics for a single set of Z-residuals. 

```{r}
#| label: diagnosis-comparison
#| code-fold: true
#| fig-width: 12
#| fig-height: 8

# Change this value (1 to 10) to see other replicated plots
i <- 1

# Set up 2 rows (Wrong vs Correct) and 3 columns (QQ, Scatter, Box)
par(mfrow = c(2, 3), mar = c(4, 4, 3, 1))

# --- ROW 1: WRONG MODEL (Linear) ---
# 1. Q-Q Plot
qqnorm(z_wrong, irep = i, main = "Wrong: Z-Resid Q-Q")
qqline(as.vector(z_wrong[, i]), col = "red")

# 2. Scatter Plot (Note the wave pattern in the residuals)
plot(z_wrong, x_axis_var = "lp", category = dat$y, 
     irep = i, add_lowess = TRUE, main = "Wrong: Z-Resid vs LP")

# 3. Boxplot
boxplot(z_wrong, x_axis_var = "x", num.bin = 10, 
        irep = i, main = "Wrong: Z-Resid vs X")

# --- ROW 2: CORRECT MODEL (Non-linear) ---
# 1. Q-Q Plot
qqnorm(z_correct, irep = i, main = "Correct: Z-Resid Q-Q")
qqline(as.vector(z_correct[, i]), col = "blue")

# 2. Scatter Plot (Residuals look like random noise)
plot(z_correct, x_axis_var = "lp", category = dat$y, 
     irep = i, add_lowess = TRUE, main = "Correct: Z-Resid vs LP")

# 3. Boxplot
boxplot(z_correct, x_axis_var = "x", num.bin = 10, 
        irep = i, main = "Correct: Z-Resid vs X")
```

> **Note:** To inspect different realizations of the randomized residuals, change the value of `i` (the `irep` index) in the code chunk above.

**Replicated p-values of the same fitted models**
```{r}
#| label: replicated-p-values-spanned
#| code-fold: true

library(knitr)
library(kableExtra)

# 1. Create the data frame 
res_table <- data.frame(
  "Replicate" = paste("Rep", 1:ncol(z_wrong)),
  "SW_W"      = as.numeric(sw.test.zresid(z_wrong)),
  "AOV_W"     = as.numeric(aov.test.zresid(z_wrong, X = "x")),
  "BL_W"      = as.numeric(bartlett.test.zresid(z_wrong, X = "x")),
  "SW_C"      = as.numeric(sw.test.zresid(z_correct)),
  "AOV_C"     = as.numeric(aov.test.zresid(z_correct, X = "x")),
  "BL_C"      = as.numeric(bartlett.test.zresid(z_correct, X = "x"))
)

# 2. Render with minimalistic styling and spanning headers
res_table %>%
  kable(
    digits = 4, 
    # Base column names (the bottom row of the header)
    col.names = c("Replicate", "SW", "AOV", "Bartlett", "SW", "AOV", "Bartlett"),
    caption = "Diagnostic Test p-values Across Randomized Replicates",
    align = "lcccccc"
  ) %>%
  kable_styling(
    bootstrap_options = c("striped", "condensed"), 
    full_width = FALSE,
    position = "center"
  ) %>%
  # Add the spanning header (the top row)
  add_header_above(c(" " = 1, "Wrong Model (Linear)" = 3, "Correct Model (Non-linear)" = 3))
```

## 4. Sampling Distribution of AOV Nonlinearity Tests

We run a simulation loop to verify that the built-in implementation correctly handles the binomial data and produces appropriate $p$-values.

```{r}
#| label: replication-verify-anova
#| code-fold: true

n_sim <- 100
sim_file <- paste0(file_prefix, "simulation_results.rds")

if (!force_rerun && file.exists(sim_file)) {
  sim_res <- readRDS(sim_file)
  p_c <- sim_res$p_c
  p_w <- sim_res$p_w
} else {
  p_c <- numeric(n_sim)
  p_w <- numeric(n_sim)
  n_sample <- 200 

  for(i in 1:n_sim) {
    x_sim <- runif(n_sample, 0, 10)
    
    # Generate true non-linear data
    eta_sim <- 2.5 * sin(x_sim)
    y_sim   <- rbinom(n_sample, 1, plogis(eta_sim))
    dat_sim <- data.frame(y = y_sim, x = x_sim)
    
    # --- Case 1: Wrong Model (Linear) ---
    fit_w <- glm(y ~ x, family = binomial(), data = dat_sim)
    z_w   <- Zresidual(fit_w, data = dat_sim, randomized = TRUE, nrep = 1)
    p_w[i] <- as.numeric(aov.test.zresid(z_w, X = "x"))
    
    # --- Case 2: Correct Model (Sine) ---
    fit_c <- glm(y ~ sin(x), family = binomial(), data = dat_sim)
    z_c   <- Zresidual(fit_c, data = dat_sim, randomized = TRUE, nrep = 1)
    p_c[i] <- as.numeric(aov.test.zresid(z_c, X = "x"))
  }
  
  saveRDS(list(p_c = p_c, p_w = p_w), sim_file)
}
```

```{r}
#| label: replication-hists
#| fig-width: 10
#| fig-height: 5
#| code-fold: true

par(mfrow = c(1, 2))
hist(p_c, main = "Correct: Uniform p-values", xlab = "p-value", col = "lightblue", breaks = 20)
abline(h = n_sim / 20, col = "red", lty = 2)

hist(p_w, main = "Wrong: Small p-values", xlab = "p-value", col = "salmon", breaks = 20)
```

**Discussion:** For the **Correct Model**, the distribution of $p$-values is uniform, indicating a well-specified model. For the **Wrong Model**, most $p$-values are small (clustered near 0), indicating that the ANOVA test successfully detected the non-linear trend.

