---
title: "Z-residuals Diagnostics for Cox Proportional Hazards Models"
subtitle: "Checking Functional Forms and Proportional Hazards Assumptions"
author: "Tingxuan Wu and Longhai Li"
format:
  html:
    code-fold: true
    message: false
vignette: >
  %\VignetteIndexEntry{Z-residual diagnostics for Cox proportional hazards models}
  %\VignetteEncoding{UTF-8}
  %\VignetteEngine{knitr::rmarkdown}
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(
  fig.width = 7,
  fig.height = 5,
  fig.align = "center"
)

set.seed(123)
force_rerun <- FALSE
```

```{r setup-paths}
#| include: false

# Define the local resources path inside the vignettes directory
resources_path <- "resources/coxph_nphnl_results"

# Ensure the directory exists
if (!dir.exists(resources_path)) {
  dir.create(resources_path, recursive = TRUE)
}
```

## Installing and Loading Required Packages

**Install Zresidual**
```{r install-zresidual}
#| eval: false
#| output: false
#| message: false
if (!requireNamespace("Zresidual", quietly = TRUE)) {
  if (!requireNamespace("remotes", quietly = TRUE)) install.packages("remotes")
  remotes::install_github("tiw150/Zresidual", upgrade = "never", dependencies = TRUE)
}
```

**Loading Required Packages**
```{r load-packages}
#| output: false
#| message: false
#| warning: false

library(Zresidual)

pkgs <- c(
  "survival", "EnvStats", "dplyr", "tibble", "gt", "gifski","MASS"
)

missing_pkgs <- pkgs[!vapply(pkgs, requireNamespace, logical(1), quietly = TRUE)]
if (length(missing_pkgs)) {
  install.packages(missing_pkgs, dependencies = TRUE)
}

invisible(lapply(pkgs, function(p) {
  suppressPackageStartupMessages(library(p, character.only = TRUE))
}))
```

**Additional Helper Functions**

```{r}
# Lightweight cumulative martingale comparator used in the original analysis script.
lwy_cum_martingale <- function(fit, x, B = 1000, seed = 1) {
  r <- residuals(fit, type = "martingale")
  ok <- is.finite(r) & is.finite(x)
  r  <- r[ok]
  x  <- x[ok]

  ord   <- order(x)
  x_ord <- x[ord]
  r_ord <- r[ord]
  n     <- length(r_ord)

  cum_obs <- cumsum(r_ord) / sqrt(n)
  stat_obs <- max(abs(cum_obs))

  set.seed(seed)
  boot_mat  <- matrix(NA_real_, nrow = B, ncol = n)
  boot_stat <- numeric(B)

  for (b in seq_len(B)) {
    w <- stats::rnorm(n)
    cum_b <- cumsum(r_ord * w) / sqrt(n)
    boot_mat[b, ] <- cum_b
    boot_stat[b]  <- max(abs(cum_b))
  }

  list(
    x = x_ord,
    cum = cum_obs,
    low = apply(boot_mat, 2, stats::quantile, probs = 0.025, na.rm = TRUE),
    high = apply(boot_mat, 2, stats::quantile, probs = 0.975, na.rm = TRUE),
    p = mean(boot_stat >= stat_obs)
  )
}
```

## Introduction

This vignette demonstrates how to use the `Zresidual` package to reproduce the core workflow of the Cox proportional hazards paper:

1. compare competing functional forms in a real-data example,
2. use joint interpretation of **Z-SW** and **Z-AOV-LP**,
3. localize lack of fit with covariate-specific diagnostics, and
4. illustrate the contrast between **functional-form misspecification** and **time-varying effects** through simulation.

The practical interpretation is:

- **non-significant Z-SW + significant Z-AOV-LP**: more consistent with functional-form misspecification,
- **significant Z-SW + non-significant Z-AOV-LP**: more consistent with non-proportional hazards,
- **both significant**: broader inadequacy,
- **neither significant**: no strong evidence for immediate revision.


## A Real-Data Example: PBC Example

### Prepare the Data

We use the Mayo Clinic PBC data from the `survival` package. Following the paper, transplantation is treated as right censoring for the cause-specific hazard of death.

```{r}
#| code-fold: false
pbc2 <- survival::pbc
pbc2$event   <- as.integer(pbc2$status == 2)
pbc2$edema   <- as.factor(pbc2$edema)
pbc2$logbili <- log(pbc2$bili)

vars_needed <- c("time", "event", "age", "edema", "bili", "logbili", "albumin")
pbc2 <- pbc2[complete.cases(pbc2[, vars_needed]), ]
```

### Fit Linear and Log-Bilirubin Cox Models

```{r}
#| code-fold: false
fit_pbc_linear <- coxph(Surv(time, event) ~ age + edema + bili + albumin, data = pbc2)
fit_pbc_log    <- coxph(Surv(time, event) ~ age + edema + logbili + albumin, data = pbc2)

AIC(fit_pbc_linear)
AIC(fit_pbc_log)

score_pbc_linear <- cox.zph(fit_pbc_linear, transform = "identity")
score_pbc_log    <- cox.zph(fit_pbc_log, transform = "identity")
```

### Compute Z-Residuals

```{r}
#| label: pbc-fitting
#| code-fold: false
#| message: true
Z_pbc_linear <- Zresidual(fit = fit_pbc_linear, data = pbc2, nrep = 500)
Z_pbc_log    <- Zresidual(fit = fit_pbc_log, data = pbc2, nrep = 500)
zcov_pbc_linear    <- Zcov(fit_pbc_linear,    data = pbc2)
zcov_pbc_log <- Zcov(fit_pbc_log, data = pbc2)
```

### Diagnostic Plots for Overall Goodness-of-fit

```{r fig-pbc-combined-animation}
#| fig-cap: "Animated Z-residual diagnostics for the PBC models across 10 randomization replicates. Top row: Linear bilirubin model. Bottom row: Log-bilirubin model. Columns (left to right): Q-Q plot, Scatterplot vs. LP, Boxplot vs. LP."
#| code-summary: "GIF generation code (folded)"
#| message: false
#| code-fold: true

pbc_combined_path <- file.path(resources_path, "pbc_combined_anim.gif")

if (force_rerun || !file.exists(pbc_combined_path)) {
  gifski::save_gif(
    expr = {
      for (i in 1:10) {
        # Set up a 2x3 grid: 2 rows (models), 3 columns (plot types)
        par(mfrow = c(2, 3), mar = c(4, 4, 3, 2))
        
        # Row 1: Linear Bilirubin Model
        qqnorm(Z_pbc_linear, irep = i, main = "Q-Q: PBC linear")
        plot(Z_pbc_linear, info = zcov_pbc_linear, x_axis_var = "lp", irep = i, add_lowess = TRUE,
             main.title = "Scatter: PBC linear")
        boxplot(Z_pbc_linear, info = zcov_pbc_linear, x_axis_var = "lp", irep = i,
                main.title = "Boxplot: PBC linear")
        
        # Row 2: Log-Bilirubin Model
        qqnorm(Z_pbc_log, irep = i, main = "Q-Q: PBC log-bili")
        plot(Z_pbc_log, info = zcov_pbc_log, x_axis_var = "lp", irep = i, add_lowess = TRUE,
             main.title = "Scatter: PBC log-bili")
        boxplot(Z_pbc_log, info = zcov_pbc_log, x_axis_var = "lp", irep = i,
                main.title = "Boxplot: PBC log-bili")
      }
    },
    gif_file = pbc_combined_path,
    width = 1350,   
    height = 900,
    delay = 1,
    res = 96
  )
}

knitr::include_graphics(pbc_combined_path)
```

### Covariate-Specific Bilirubin Diagnostics

```{r fig-pbc-bili-animation}
#| fig-cap: "Animated bilirubin-specific Z-residual diagnostics for the PBC models. Left column: linear bilirubin model. Right column: log-bilirubin model."
#| code-summary: "GIF generation code (folded)"
#| code-fold: true
#| message: false

pbc_bili_path <- file.path(resources_path, "pbc_bili_anim.gif")

if (force_rerun || !file.exists(pbc_bili_path)) {
  gifski::save_gif(
    expr = {
      for (i in 1:10) {
        par(mfrow = c(2, 2), mar = c(4, 4, 2, 2))
        plot(Z_pbc_linear,info = zcov_pbc_linear, x_axis_var = "bili", irep = i, add_lowess = TRUE,
             main.title = "Scatter: linear bili")
        plot(Z_pbc_log,info = zcov_pbc_log,  x_axis_var = "logbili", irep = i, add_lowess = TRUE,
             main.title = "Scatter: log(bili)")
        boxplot(Z_pbc_linear, info = zcov_pbc_linear, x_axis_var = "bili", irep = i,
                main.title = "Boxplot: linear bili")
        boxplot(Z_pbc_log, info = zcov_pbc_log, x_axis_var = "logbili", irep = i,
                main.title = "Boxplot: log(bili)")
      }
    },
    gif_file = pbc_bili_path,
    width = 900,
    height = 900,
    delay = 1,
    res = 96
  )
}

knitr::include_graphics(pbc_bili_path)
```


### Replicated P-values with Z-residual Diagnostics of the Same Models

```{r include=FALSE}
# Compute replicated p-values and diagnostics before plotting
sw_pbc_linear      <- sw.test.zresid(Z_pbc_linear)
sw_pbc_log         <- sw.test.zresid(Z_pbc_log)
aovlp_pbc_linear   <- aov.test.zresid(Z_pbc_linear, X = "lp", zcov = zcov_pbc_linear, k.anova = 10)
aovlp_pbc_log      <- aov.test.zresid(Z_pbc_log, X = "lp", zcov= zcov_pbc_log, k.anova = 10)
aovbili_pbc_linear <- aov.test.zresid(Z_pbc_linear, X = "bili", zcov = zcov_pbc_linear, k.anova = 10)
aovbili_pbc_log    <- aov.test.zresid(Z_pbc_log, X = "logbili", zcov= zcov_pbc_log, k.anova = 10)

lwy_pbc_linear     <- lwy_cum_martingale(fit_pbc_linear, x = pbc2$bili, B = 1000, seed = 1)
lwy_pbc_log        <- lwy_cum_martingale(fit_pbc_log, x = pbc2$logbili, B = 1000, seed = 1)
```

**Histograms of Replicated P-values**

```{r fig-pbc-pvalue-hist}
#| fig-height: 8
#| fig-width: 9
#| fig-cap: "Histograms of replicated Z-residual test p-values for the PBC models. Red lines indicate the conservative upper-bound p-values based on the median order statistic."

par(mfrow = c(3, 2), mar = c(4, 4, 2, 2))

hist(sw_pbc_linear, main = "Linear: Z-SW", breaks = 20)
abline(v = upper_bound_pvalue(sw_pbc_linear), col = "red", lwd = 2)

hist(sw_pbc_log, main = "Log: Z-SW", breaks = 20)
abline(v = upper_bound_pvalue(sw_pbc_log), col = "red", lwd = 2)

hist(aovlp_pbc_linear, main = "Linear: Z-AOV-LP", breaks = 20)
abline(v = upper_bound_pvalue(aovlp_pbc_linear), col = "red", lwd = 2)

hist(aovlp_pbc_log, main = "Log: Z-AOV-LP", breaks = 20)
abline(v = upper_bound_pvalue(aovlp_pbc_log), col = "red", lwd = 2)

hist(aovbili_pbc_linear, main = "Linear: Z-AOV-bili", breaks = 20)
abline(v = upper_bound_pvalue(aovbili_pbc_linear), col = "red", lwd = 2)

hist(aovbili_pbc_log, main = "Log: Z-AOV-log(bili)", breaks = 20)
abline(v = upper_bound_pvalue(aovbili_pbc_log), col = "red", lwd = 2)
```

**Summary of Replicated P-values**
```{r}
pbc_summary <- tibble::tibble(
  Model = c("Linear bili", "Log(bili)"),
  AIC = c(AIC(fit_pbc_linear), AIC(fit_pbc_log)),
  `Score global` = c(score_pbc_linear$table["GLOBAL", "p"], score_pbc_log$table["GLOBAL", "p"]),
  `Score bili` = c(score_pbc_linear$table["bili", "p"], score_pbc_log$table["logbili", "p"]),
  `LWY-like bili` = c(lwy_pbc_linear$p, lwy_pbc_log$p),
  `Z-SW` = c(upper_bound_pvalue(sw_pbc_linear), upper_bound_pvalue(sw_pbc_log)),
  `Z-AOV-LP` = c(upper_bound_pvalue(aovlp_pbc_linear), upper_bound_pvalue(aovlp_pbc_log)),
  `Z-AOV-bili` = c(upper_bound_pvalue(aovbili_pbc_linear), upper_bound_pvalue(aovbili_pbc_log))
)

pbc_summary %>%
  gt::gt() %>%
  gt::fmt_number(columns = -Model, decimals = 3) %>%
  gt::tab_header(title = "PBC model comparison")
```

## A Simulation Study with Non-linear Cox PH Datasets


**Data Generation:**

In this setting, the data are generated under a proportional-hazards model with a **log(x2)** effect, but we first fit a misspecified model using raw `x2`.

```{r}

# Functions for Simulating Datasets

rexp2 <- function(n, rate){ if (rate==0) rep(Inf,n) else rexp(n = n, rate = rate)}

## Genearting Non-linear Cox PH Data

sim_nonlinear_coxph <- function(
  sample_size = 500,
  lambda = 0.007,
  alpha = 3,
  mean.censor = 7.5,
  beta_a=-2,
  beta_b=-2,
  beta1 = -0.5,
  corr = 0,
  t0=pi/2,
  seed = 1
) {
  set.seed(seed)
  n <- sample_size
  Sigma <- matrix(c(2, corr, corr, 2), 2, 2)
  X <- MASS::mvrnorm(n = n, mu = c(0, 0), Sigma = Sigma)

  x1 <- round(X[, 1], 1)
  x2 <- pmax(abs(round(X[, 2], 1)), 0.01)
  logx2 <- log(x2)

  v <- runif(n)
  a<- -log(v)
  b<- lambda*exp(x1*beta1+beta_a*log(x2)) * (t0^alpha)
  surv_time<-rep(0,n)
  group1<-which(a<b)
  group2<-which(a>=b)
  surv_time[group1]<- (-log(v)[group1]/
                         (lambda*exp(x1[group1]*beta1+log(x2)[group1]*beta_a)))^(1 /alpha)
  
  surv_time[group2]<- ((-log(v)[group2]-
                          ((t0^alpha)*lambda*
                             exp(x1[group2]*beta1+log(x2)[group2]*beta_a))+
                          ((t0^alpha)*lambda*
                             exp(x1[group2]*beta1+log(x2)[group2]*beta_b)))
                       /(lambda*exp(x1[group2]*beta1)
                         *exp(log(x2)[group2]*beta_b)))^(1 /alpha)
  # Censoring times
  c0<- rexp2(n, rate= 1/mean.censor)
  data.frame(
    x1 = x1,
    x2 = x2,
    logx2 = logx2,
    t = pmin(surv_time, c0),
    d = as.integer(surv_time <= c0)
  )
}

## Genearting Time Varying Cox PH Data
sim_timevarying_coxph <- function(
  sample_size = 500,
  lambda = 0.007,
  alpha = 3,
  mean.censor = 7.5,
  beta1 = -2,
  beta_a = 1.35,
  beta_b = -1.35,
  t0 = pi / 2,
  corr = 0,
  seed = 1
) {
  set.seed(seed)
  n <- sample_size
  sigma<-rbind(c(2,corr),c(corr,2))
  mu<-c(0,0) 
  covar<-MASS::mvrnorm(n=n, mu=mu, Sigma=sigma)
  x1 = round(covar[,1],1)
  x2 = round(covar[,2],1)

  v <- runif(n)
  a<- -log(v)
  b<- lambda*exp(x1*beta1+beta_a*x2) * (t0^alpha)
  surv_time<-rep(0,n)
  group1<-which(a<b)
  group2<-which(a>=b)
  surv_time[group1]<- (-log(v)[group1]/
                         (lambda*exp(x1[group1]*beta1+x2[group1]*beta_a)))^(1 /alpha)
  
  surv_time[group2]<- ((-log(v)[group2]-
                          ((t0^alpha)*lambda*
                             exp(x1[group2]*beta1+x2[group2]*beta_a))+
                          ((t0^alpha)*lambda*
                             exp(x1[group2]*beta1+x2[group2]*beta_b)))
                       /(lambda*exp(x1[group2]*beta1)
                         *exp(x2[group2]*beta_b)))^(1 /alpha)
  # Censoring times
  c0<- rexp2(n, rate= 1/mean.censor)
  
  data.frame(
    x1 = x1,
    x2 = x2,
    t = pmin(surv_time, c0),
    d = as.integer(surv_time <= c0)
  )
}
```

### Single Dataset Evaluation

```{r}
dat_nonlinear <- sim_nonlinear_coxph(seed = 101)

fit_linear <- coxph(Surv(t, d) ~ x1 + x2, data = dat_nonlinear)
fit_nonlinear <- coxph(Surv(t, d) ~ x1 + log(x2), data = dat_nonlinear)

score_linear <- cox.zph(fit_linear, transform = "identity")
score_nonlinear <- cox.zph(fit_nonlinear, transform = "identity")

Z_linear <- Zresidual(fit = fit_linear, data = dat_nonlinear, nrep = 200)
Z_nonlinear <- Zresidual(fit = fit_nonlinear, data = dat_nonlinear, nrep = 200)

zcov_linear    <- Zcov(fit_linear,    data = dat_nonlinear)
zcov_nonlinear <- Zcov(fit_nonlinear, data = dat_nonlinear)
```

```{r fig-sim-nonlinear}
#| fig-cap: "A comprehensive diagnostic analysis using Z-residuals from simulated data. The upper row illustrates model diagnostics for a poorly specified linear-form fit. The lower row shows diagnostics for a correctly specified nonlinear model incorporating log-transformed covariates."
#| fig-width: 16
#| fig-height: 10
#| out-width: 100%
#| dpi: 300
#| message: false
#| results: false

par(mfrow = c(2, 3), mar = c(4, 4, 3, 3), cex = 1.1)

# Row 1: Misspecified Linear Fit
qqnorm(Z_linear, irep = 1, main = "Linear Model Q-Q Z-Residuals")
plot(Z_linear, info=zcov_linear, x_axis_var = "lp", irep = 1, add_lowess = TRUE, 
     main.title = "Linear Model Z-Residuals vs LP")
plot(Z_linear, info=zcov_linear, x_axis_var = "x2", irep = 1, add_lowess = TRUE, 
     main.title = "Linear Model Z-Residuals vs x2")

# Row 2: Correctly Specified Nonlinear Fit
qqnorm(Z_nonlinear, irep = 1, main = "Nonlinear Model Q-Q Z-Residuals")
plot(Z_nonlinear, info=zcov_nonlinear, x_axis_var = "lp", irep = 1, add_lowess = TRUE, 
     main.title = "Nonlinear Model Z-Residuals vs LP")
plot(Z_nonlinear, info=zcov_nonlinear, x_axis_var = "log(x2)", irep = 1, add_lowess = TRUE, 
     main.title = "Nonlinear Model Z-Residuals vs log(x2)")
```

```{r}
sw_linear    <- sw.test.zresid(Z_linear)
sw_nonlinear     <- sw.test.zresid(Z_nonlinear)
aovlp_linear <- aov.test.zresid(Z_linear, X = "lp",zcov = zcov_linear, k.anova = 10)
aovlp_nonlinear  <- aov.test.zresid(Z_nonlinear, X = "lp",zcov = zcov_nonlinear, k.anova = 10)
aovx_linear  <- aov.test.zresid(Z_linear, X = "x2",zcov = zcov_linear, k.anova = 10)
aovx_nonlinear   <- aov.test.zresid(Z_nonlinear, X = "log(x2)",zcov = zcov_nonlinear, k.anova = 10)

sim_linear_summary <- tibble::tibble(
  Model = c("Linear Cox-PH", "Nonlinear Cox-PH"),
  `Score global` = c(score_linear$table["GLOBAL", "p"], score_nonlinear$table["GLOBAL", "p"]),
  `Score target covariate` = c(score_linear$table["x2", "p"], score_nonlinear$table["log(x2)", "p"]),
  `Z-SW` = c(upper_bound_pvalue(sw_linear), upper_bound_pvalue(sw_nonlinear)),
  `Z-AOV-LP` = c(upper_bound_pvalue(aovlp_linear), upper_bound_pvalue(aovlp_nonlinear)),
  `Z-AOV-target` = c(upper_bound_pvalue(aovx_linear), upper_bound_pvalue(aovx_nonlinear))
)

sim_linear_summary %>%
  gt::gt() %>%
  gt::fmt_number(columns = -Model, decimals = 3) %>%
  gt::tab_header(title = "Summary of Replicated P-values for the Nonlinear Dataset")
```

### Checking the Sampling Distribution of Shapiro-Wilk Tests p-values

We perform a simulation loop across 500 datasets to compare the $p$-values produced by the Shapiro-Wilk diagnostic test for both the misspecified linear model and the correctly specified log-nonlinear model. We save these computation results to avoid repeating them in future renders unless `force_rerun` is enabled.

```{r replication-verify-sw}
#| code-fold: true
#| results: hide
#| message: false
#| warning: false

n_sim <- 500
sim_file <- file.path(resources_path, "sim_sw_pvalues.rds")

if (!force_rerun && file.exists(sim_file)) {
  # Load pre-computed results
  sim_res <- readRDS(sim_file)
  p_c <- sim_res$p_c
  p_w <- sim_res$p_w
} else {
  # Execute the simulation loop
  p_c <- numeric(n_sim)
  p_w <- numeric(n_sim)
  
  for(i in 1:n_sim) {
    # Generate true non-linear data
    dat_sim <- sim_nonlinear_coxph(seed = i)
    
    # --- Case 1: Wrong Model (Linear) ---
    fit_w_sim <- coxph(Surv(t, d) ~ x1 + x2, data = dat_sim)
    z_w_sim   <- Zresidual(fit_w_sim, data = dat_sim, nrep = 1)
    
    p_w[i] <- as.numeric(sw.test.zresid(z_w_sim))
    
    # --- Case 2: Correct Model (Nonlinear) ---
    fit_c_sim <- coxph(Surv(t, d) ~ x1 + log(x2), data = dat_sim)
    z_c_sim   <- Zresidual(fit_c_sim, data = dat_sim, nrep = 1)
    
    p_c[i] <- as.numeric(sw.test.zresid(z_c_sim))
  }
  
  # Save the results for future renders
  saveRDS(list(p_c = p_c, p_w = p_w), sim_file)
}
```

```{r replication-hists-sw}
#| fig-width: 10
#| fig-height: 5
#| code-fold: true
#| fig-cap: "Distributions of Shapiro-Wilk diagnostic test p-values across 500 simulated datasets."

par(mfrow = c(1, 2))
hist(p_c, main = "Correct: Uniform p-values", xlab = "p-value", col = "lightblue", breaks = 10)
abline(h = n_sim / 10, col = "red", lty = 2)

hist(p_w, main = "Wrong: Small p-values", xlab = "p-value", col = "salmon", breaks = 10)
```

**Discussion:** For the **Correct Model** (log-nonlinear), the distribution of $p$-values is relatively uniform, indicating a well-specified functional form that yields normally distributed Z-residuals. For the **Wrong Model** (linear), the vast majority of $p$-values are clustered near 0, demonstrating that the Shapiro-Wilk test reliably detects the non-normality caused by the misspecified functional form.
