---
title: "Z-Residual Diagnostic Tool for Assessing Covariate Functional Form"
author: "Tingxuan Wu"
date: "2026-07-30"
format:
  html:
    toc: true
    toc-depth: 3
    code-fold: true
execute:
  message: false
  warning: false
vignette: >
  %\VignetteIndexEntry{Z-residual diagnostic tool for assessing covariate functional form}
  %\VignetteEncoding{UTF-8}
  %\VignetteEngine{quarto::html}
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(
  fig.width = 7,
  fig.height = 5,
  fig.align = "center",
  warning = FALSE
)

set.seed(123)

# Number of randomized Z-residual replicates used for inference.
nrep <- 1000L

# The animations use only the first few replicates so that rendering remains
# reasonably fast. Set force_rerun to TRUE to rebuild saved animations.
animation_reps <- 10L
force_rerun <- FALSE
```

```{r setup-paths}
#| include: false

resources_path <- "resources/demo_zresid_frailty_model_results"

if (!dir.exists(resources_path)) {
  dir.create(resources_path, recursive = TRUE)
}

make_zresidual_gif <- function(
  output_path,
  nframes,
  width,
  height,
  draw_frame,
  delay = 0.75
) {
  if (!force_rerun && file.exists(output_path)) {
    return(output_path)
  }

  if (!requireNamespace("gifski", quietly = TRUE)) {
    stop(
      "The 'gifski' package is required to build the animated figures.",
      call. = FALSE
    )
  }

  frame_dir <- tempfile("zresidual_frames_")
  dir.create(frame_dir)
  on.exit(unlink(frame_dir, recursive = TRUE, force = TRUE), add = TRUE)

  frame_files <- file.path(
    frame_dir,
    sprintf("frame_%03d.png", seq_len(nframes))
  )

  for (i in seq_len(nframes)) {
    grDevices::png(
      filename = frame_files[i],
      width = width,
      height = height,
      res = 120
    )

    tryCatch(
      draw_frame(i),
      finally = grDevices::dev.off()
    )
  }

  unlink(output_path, force = TRUE)

  gifski::gifski(
    png_files = frame_files,
    gif_file = output_path,
    width = width,
    height = height,
    delay = delay,
    loop = TRUE
  )

  output_path
}
```

## Introduction

This vignette explains how to use the `Zresidual` package to calculate
Z-residuals from models fitted with `coxph()` in the `survival` package. It
also demonstrates how Z-residuals can assess overall goodness of fit (GOF) and
identify specific model misspecifications in semiparametric shared-frailty
models.

For detailed definitions and the complete data analysis, refer to the paper
*Z-residual diagnostics for detecting misspecification of the functional form
of covariates for shared frailty models*.

## Definition of Z-residual

We use Z-residuals to diagnose shared-frailty models in a Cox proportional
hazards setting with an unspecified baseline hazard. Suppose there are $g$
groups of individuals, with group $i$ containing $n_i$ individuals,
$i = 1,\ldots,g$. Let $y_{ij}$ be a possibly right-censored observation for
the $j$th individual in group $i$, and let $\delta_{ij}$ indicate whether the
observation is uncensored.

The randomized survival probability (RSP) for $y_{ij}$ is

$$
S_{ij}^{R}(y_{ij}, \delta_{ij}, U_{ij}) =
\begin{cases}
S_{ij}(y_{ij}),
  & \text{if } \delta_{ij}=1,\\
U_{ij}S_{ij}(y_{ij}),
  & \text{if } \delta_{ij}=0,
\end{cases}
$$

where $U_{ij}$ is a uniform random variable on $(0,1)$ and
$S_{ij}(\cdot)$ is the postulated survival function given the covariates
$x_{ij}$. When $y_{ij}$ is censored, $S_{ij}^{R}$ is randomized between 0
and $S_{ij}(y_{ij})$.

Under the true model, the RSPs are uniformly distributed on $(0,1)$ given
$x_i$. They can therefore be transformed into residuals with a desired
reference distribution. We use the standard normal quantile:

$$
r_{ij}^{Z}(y_{ij}, \delta_{ij}, U_{ij})
= -\Phi^{-1}\left\{
S_{ij}^{R}(y_{ij}, \delta_{ij}, U_{ij})
\right\}.
$$

Under the true model, these residuals follow a standard normal distribution.
Consequently, diagnostic methods developed for normal regression can also be
used with censored data. The normal transformation also makes extremely small
or large RSPs easier to identify; even a small number of such extreme values
may indicate model misspecification.

## Installation

You can install the development version of `Zresidual` from GitHub:

```{r install-zresidual}
#| eval: false
#| output: false
#| message: false
if (!requireNamespace("Zresidual", quietly = TRUE)) {
  if (!requireNamespace("remotes", quietly = TRUE)) install.packages("remotes")
  remotes::install_github("tiw150/Zresidual", upgrade = "never", dependencies = TRUE)
}
```

Load the packages used in this vignette:

```{r load-packages}
#| output: false
#| message: false
#| warning: false

library(Zresidual)
library(survival)
```

The optional `DT` and `gifski` packages are used for interactive tables and
animated diagnostic figures, respectively.


## Examples for Illustration and Demonstration

### A real dataset

This example illustrates how Z-residuals diagnose both overall GOF and the
functional form of covariates in an analysis of survival times for patients
with acute myeloid leukemia.

The dataset contains 411 patients recorded at the M. D. Anderson Cancer Center
between 1980 and 1996. The patients were under age 60 and came from 24
administrative districts. The available prognostic factors include age, sex,
white blood cell count (`wbc`) at diagnosis, and the Townsend score (`tpi`),
where higher values indicate less affluent areas. The censoring rate is 29.2%.
The response is survival time in days from study entry to death or censoring.

Preliminary analysis showed that `wbc` is highly right-skewed. A logarithmic
transformation is often used to reduce the influence of extremely large
covariate values, but it may also mask their relationship with the outcome.

```{r prepare-data}
#| code-fold: false

leuksurv_path <- system.file(
  "extdata",
  "LeukSurv.rda",
  package = "Zresidual"
)

if (nzchar(leuksurv_path)) {
  load(leuksurv_path)
} else {
  # Compatibility with package versions that exposed the dataset through data().
  data("LeukSurv", package = "Zresidual", envir = environment())
}

if (!exists("LeukSurv")) {
  stop("The LeukSurv dataset could not be loaded.", call. = FALSE)
}

LeukSurv$district <- as.factor(LeukSurv$district)
LeukSurv$sex <- as.factor(LeukSurv$sex)
LeukSurv$logwbc <- log(LeukSurv$wbc + 0.001)
LeukSurv <- droplevels(LeukSurv[LeukSurv$age < 60, ])
```

```{r table-leuksurv}
#| code-fold: true

if (requireNamespace("DT", quietly = TRUE)) {
  DT::datatable(
    LeukSurv,
    caption = "Table 1: Acute myeloid leukemia survival data.",
    options = list(pageLength = 10, scrollX = TRUE)
  )
} else {
  knitr::kable(
    head(LeukSurv, 20),
    caption = "Table 1: Acute myeloid leukemia survival data (first 20 rows)."
  )
}
```

### Fit the models

We fit two shared-frailty models. The **wbc model** includes untransformed
`wbc`, age, sex, and `tpi`. The **lwbc model** replaces `wbc` with
`log(wbc)`.

```{r fit-models}
#| code-fold: false

fit_LeukSurv_wbc <- coxph(
  Surv(time, cens) ~ age + sex + wbc + tpi +
    frailty(district, distribution = "gamma"),
  data = LeukSurv
)

fit_LeukSurv_logwbc <- coxph(
  Surv(time, cens) ~ age + sex + logwbc + tpi +
    frailty(district, distribution = "gamma"),
  data = LeukSurv
)
```

### Z-residual calculation

Once the models have been fitted, calculate 1,000 randomized Z-residual
replicates and extract the aligned covariate information used by the plotting
and testing functions.

```{r calculate-zresiduals}
#| code-fold: false
#| message: true

Zresid.LeukSurv.wbc <- Zresidual(
  fit = fit_LeukSurv_wbc,
  data = LeukSurv,
  nrep = nrep,
  seed = 123
)

Zresid.LeukSurv.logwbc <- Zresidual(
  fit = fit_LeukSurv_logwbc,
  data = LeukSurv,
  nrep = nrep,
  seed = 456
)

zcov.LeukSurv.wbc <- Zcov(fit_LeukSurv_wbc, data = LeukSurv)
zcov.LeukSurv.logwbc <- Zcov(fit_LeukSurv_logwbc, data = LeukSurv)
```

### Graphically assess the models

#### Overall goodness of fit

A normal Q-Q plot of the Z-residuals provides a graphical assessment of
overall GOF. The Shapiro-Wilk (SW) and Shapiro-Francia (SF) normality tests
provide numerical assessments. For these data, the Q-Q plots for both models
align reasonably well with the 45-degree reference line, and the Z-SW tests
generally give large p-values.

::: {#fig-zresidual-qq-animation layout-ncol=2}

```{r zresidual-qq-wbc-animation}
#| code-summary: "wbc model code"
#| code-fold: true
#| message: false
#| fig-show: animate
#| animation-hook: gifski
#| interval: 0.75
#| fig-width: 6
#| fig-height: 5

for (i in 1:animation_reps) {
  qqnorm(
    Zresid.LeukSurv.wbc,
    zcov = zcov.LeukSurv.wbc,
    irep = i,
    main.title = "Z-residual Q-Q plot: wbc model"
  )
}
```

```{r zresidual-qq-lwbc-animation}
#| code-summary: "lwbc model code"
#| code-fold: true
#| message: false
#| fig-show: animate
#| animation-hook: gifski
#| interval: 0.75
#| fig-width: 6
#| fig-height: 5

for (i in 1:animation_reps) {
  qqnorm(
    Zresid.LeukSurv.logwbc,
    zcov = zcov.LeukSurv.logwbc,
    irep = i,
    main.title = "Z-residual Q-Q plot: lwbc model"
  )
}
```

Q-Q plots of Z-residuals for the wbc model (left) and lwbc model
(right) across randomized replicates.

:::


The Z-residuals can also be grouped by intervals of the linear predictor or a
covariate. We can then assess whether the grouped residuals have homogeneous
distributions. ANOVA F-tests compare group means, and Bartlett tests compare
group variances.

For both models, the scatterplots of Z-residuals against the linear predictor
show no prominent trend: the LOWESS curves remain close to zero. The grouped
boxplots also suggest similar means and variances. Correspondingly, Z-AOV-LP
and Z-BL-LP generally produce large p-values.

::: {#fig-zresidual-lp-animation layout-ncol=2}

```{r zresidual-lp-scatter-wbc-animation}
#| code-summary: "wbc scatterplot code"
#| code-fold: true
#| message: false
#| fig-show: animate
#| animation-hook: gifski
#| interval: 0.75
#| fig-width: 6
#| fig-height: 5

for (i in 1:animation_reps) {
  plot(
    Zresid.LeukSurv.wbc,
    zcov = zcov.LeukSurv.wbc,
    x_axis_var = "lp",
    irep = i,
    add_lowess = TRUE,
    main.title = "Scatterplot: wbc model"
  )
}
```

```{r zresidual-lp-scatter-lwbc-animation}
#| code-summary: "lwbc scatterplot code"
#| code-fold: true
#| message: false
#| fig-show: animate
#| animation-hook: gifski
#| interval: 0.75
#| fig-width: 6
#| fig-height: 5

for (i in 1:animation_reps) {
  plot(
    Zresid.LeukSurv.logwbc,
    zcov = zcov.LeukSurv.logwbc,
    x_axis_var = "lp",
    irep = i,
    add_lowess = TRUE,
    main.title = "Scatterplot: lwbc model"
  )
}
```

```{r zresidual-lp-box-wbc-animation}
#| code-summary: "wbc boxplot code"
#| code-fold: true
#| message: false
#| fig-show: animate
#| animation-hook: gifski
#| interval: 0.75
#| fig-width: 6
#| fig-height: 5

for (i in 1:animation_reps) {
  boxplot(
    Zresid.LeukSurv.wbc,
    zcov = zcov.LeukSurv.wbc,
    x_axis_var = "lp",
    irep = i,
    main.title = "Boxplot: wbc model"
  )
}
```

```{r zresidual-lp-box-lwbc-animation}
#| code-summary: "lwbc boxplot code"
#| code-fold: true
#| message: false
#| fig-show: animate
#| animation-hook: gifski
#| interval: 0.75
#| fig-width: 6
#| fig-height: 5

for (i in 1:animation_reps) {
  boxplot(
    Zresid.LeukSurv.logwbc,
    zcov = zcov.LeukSurv.logwbc,
    x_axis_var = "lp",
    irep = i,
    main.title = "Boxplot: lwbc model"
  )
}
```

Scatterplots (top row) and grouped boxplots (bottom row) of Z-residuals
against the linear predictor for the wbc model (left column) and lwbc model
(right column).

:::

#### Identifying a misspecified covariate functional form

The overall diagnostics do not reveal serious misspecification in either
model. Diagnostics against `wbc` and `logwbc`, however, suggest that the
functional form in the lwbc model is misspecified.

For the wbc model, the LOWESS curve remains close to zero and the grouped
Z-residuals have similar means and variances. For the lwbc model, the LOWESS
curve shows a nonlinear trend and the grouped residuals have different means.
The very small Z-AOV-log(wbc) p-values provide strong evidence that the log
transformation is inappropriate for modelling survival time in this example.

::: {#fig-zresidual-wbc-animation layout-ncol=2}

```{r zresidual-wbc-scatter-wbc-animation}
#| code-summary: "wbc model scatterplot code"
#| code-fold: true
#| message: false
#| fig-show: animate
#| animation-hook: gifski
#| interval: 0.75
#| fig-width: 6
#| fig-height: 5

for (i in 1:animation_reps) {
  plot(
    Zresid.LeukSurv.wbc,
    zcov = zcov.LeukSurv.wbc,
    x_axis_var = "wbc",
    irep = i,
    add_lowess = TRUE,
    main.title = "Scatterplot: wbc model"
  )
}
```

```{r zresidual-wbc-scatter-lwbc-animation}
#| code-summary: "lwbc model scatterplot code"
#| code-fold: true
#| message: false
#| fig-show: animate
#| animation-hook: gifski
#| interval: 0.75
#| fig-width: 6
#| fig-height: 5

for (i in 1:animation_reps) {
  plot(
    Zresid.LeukSurv.logwbc,
    zcov = zcov.LeukSurv.logwbc,
    x_axis_var = "logwbc",
    irep = i,
    add_lowess = TRUE,
    main.title = "Scatterplot: lwbc model"
  )
}
```

```{r zresidual-wbc-box-wbc-animation}
#| code-summary: "wbc model boxplot code"
#| code-fold: true
#| message: false
#| fig-show: animate
#| animation-hook: gifski
#| interval: 0.75
#| fig-width: 6
#| fig-height: 5

for (i in 1:animation_reps) {
  boxplot(
    Zresid.LeukSurv.wbc,
    zcov = zcov.LeukSurv.wbc,
    x_axis_var = "wbc",
    irep = i,
    main.title = "Boxplot: wbc model"
  )
}
```

```{r zresidual-wbc-box-lwbc-animation}
#| code-summary: "lwbc model boxplot code"
#| code-fold: true
#| message: false
#| fig-show: animate
#| animation-hook: gifski
#| interval: 0.75
#| fig-width: 6
#| fig-height: 5

for (i in 1:animation_reps) {
  boxplot(
    Zresid.LeukSurv.logwbc,
    zcov = zcov.LeukSurv.logwbc,
    x_axis_var = "logwbc",
    irep = i,
    main.title = "Boxplot: lwbc model"
  )
}
```

Scatterplots (top row) and grouped boxplots (bottom row) of Z-residuals
against wbc for the wbc model (left column) and log(wbc) for the lwbc model
(right column).

:::


The boxplots against the categorical covariate `sex` show similar means and
variances across groups for both models. The corresponding Z-AOV and Z-BL
p-values are generally greater than 0.05.

::: {#fig-zresidual-sex-animation layout-ncol=2}

```{r zresidual-sex-wbc-animation}
#| code-summary: "wbc model code"
#| code-fold: true
#| message: false
#| fig-show: animate
#| animation-hook: gifski
#| interval: 0.75
#| fig-width: 6
#| fig-height: 5

for (i in 1:animation_reps) {
  boxplot(
    Zresid.LeukSurv.wbc,
    zcov = zcov.LeukSurv.wbc,
    x_axis_var = "sex",
    irep = i,
    main.title = "Boxplot: wbc model"
  )
}
```

```{r zresidual-sex-lwbc-animation}
#| code-summary: "lwbc model code"
#| code-fold: true
#| message: false
#| fig-show: animate
#| animation-hook: gifski
#| interval: 0.75
#| fig-width: 6
#| fig-height: 5

for (i in 1:animation_reps) {
  boxplot(
    Zresid.LeukSurv.logwbc,
    zcov = zcov.LeukSurv.logwbc,
    x_axis_var = "sex",
    irep = i,
    main.title = "Boxplot: lwbc model"
  )
}
```

Grouped boxplots of Z-residuals against sex for the wbc model (left) and
lwbc model (right).

:::

### Quantitatively assess the models

The Shapiro-Wilk and Shapiro-Francia tests assess the normality of each
randomized Z-residual replicate and therefore the overall GOF. The
Shapiro-Francia test can also be extended to censored Z-residuals through
`gof.censored.zresidual()`.

```{r calculate-normality-tests}
#| code-fold: false

sw.wbc <- sw.test.zresid(Zresid.LeukSurv.wbc)
sw.lwbc <- sw.test.zresid(Zresid.LeukSurv.logwbc)
sf.wbc <- sf_test.zresid(Zresid.LeukSurv.wbc)
sf.lwbc <- sf_test.zresid(Zresid.LeukSurv.logwbc)

gof_tests <- data.frame(
  replicate = seq_len(nrep),
  `Z-SW: wbc model` = sw.wbc,
  `Z-SW: lwbc model` = sw.lwbc,
  `Z-SF: wbc model` = sf.wbc,
  `Z-SF: lwbc model` = sf.lwbc,
  check.names = FALSE
)
```

```{r table-normality-tests}
#| code-fold: true

if (requireNamespace("DT", quietly = TRUE)) {
  DT::datatable(
    gof_tests,
    caption = paste(
      "Table 2: Replicated Z-SW and Z-SF p-values for the wbc",
      "and lwbc models."
    ),
    options = list(pageLength = 10, scrollX = TRUE)
  )
} else {
  knitr::kable(
    head(gof_tests, 20),
    caption = paste(
      "Table 2: Replicated Z-SW and Z-SF p-values for the wbc",
      "and lwbc models (first 20 rows)."
    )
  )
}
```

The Z-residuals can be grouped by intervals of a covariate or linear
predictor. ANOVA tests compare the group means, while Bartlett tests compare
the group variances.

```{r calculate-homogeneity-tests}
#| code-fold: false

aov.wbc.lp <- aov.test.zresid(
  Zresid.LeukSurv.wbc,
  X = "lp",
  zcov = zcov.LeukSurv.wbc,
  k.anova = 10
)

aov.lwbc.lp <- aov.test.zresid(
  Zresid.LeukSurv.logwbc,
  X = "lp",
  zcov = zcov.LeukSurv.logwbc,
  k.anova = 10
)

bl.wbc.lp <- bartlett.test.zresid(
  Zresid.LeukSurv.wbc,
  X = "lp",
  zcov = zcov.LeukSurv.wbc,
  k.bl = 10
)

bl.lwbc.lp <- bartlett.test.zresid(
  Zresid.LeukSurv.logwbc,
  X = "lp",
  zcov = zcov.LeukSurv.logwbc,
  k.bl = 10
)

aov.wbc <- aov.test.zresid(
  Zresid.LeukSurv.wbc,
  X = "wbc",
  zcov = zcov.LeukSurv.wbc,
  k.anova = 10
)

aov.lwbc <- aov.test.zresid(
  Zresid.LeukSurv.logwbc,
  X = "logwbc",
  zcov = zcov.LeukSurv.logwbc,
  k.anova = 10
)

bl.wbc <- bartlett.test.zresid(
  Zresid.LeukSurv.wbc,
  X = "wbc",
  zcov = zcov.LeukSurv.wbc,
  k.bl = 10
)

bl.lwbc <- bartlett.test.zresid(
  Zresid.LeukSurv.logwbc,
  X = "logwbc",
  zcov = zcov.LeukSurv.logwbc,
  k.bl = 10
)

homogeneity_tests <- data.frame(
  replicate = seq_len(nrep),
  `Z-AOV-LP: wbc model` = aov.wbc.lp,
  `Z-AOV-LP: lwbc model` = aov.lwbc.lp,
  `Z-BL-LP: wbc model` = bl.wbc.lp,
  `Z-BL-LP: lwbc model` = bl.lwbc.lp,
  `Z-AOV-wbc: wbc model` = aov.wbc,
  `Z-AOV-log(wbc): lwbc model` = aov.lwbc,
  `Z-BL-wbc: wbc model` = bl.wbc,
  `Z-BL-log(wbc): lwbc model` = bl.lwbc,
  check.names = FALSE
)
```

```{r table-homogeneity-tests}
#| code-fold: true

if (requireNamespace("DT", quietly = TRUE)) {
  DT::datatable(
    homogeneity_tests,
    caption = paste(
      "Table 3: Replicated Z-AOV-LP, Z-AOV-covariate, Z-BL-LP,",
      "and Z-BL-covariate p-values for the wbc and lwbc models."
    ),
    options = list(pageLength = 10, scrollX = TRUE)
  )
} else {
  knitr::kable(
    head(homogeneity_tests, 20),
    caption = paste(
      "Table 3: Replicated homogeneity-test p-values",
      "(first 20 rows)."
    )
  )
}
```

The following histograms display the 1,000 replicated p-values. The red
vertical lines are the conservative upper-bound p-values obtained from the
replicated results. Z-SW, Z-SF, and Z-AOV-LP produce substantial proportions
of p-values above 0.05 for both models. In contrast, nearly all replicated
Z-AOV-log(wbc) p-values for the lwbc model are below 0.001, reinforcing the
evidence that the logarithmic transformation is inappropriate here.

```{r fig-replicated-pvalues}
#| fig-cap: "Histograms of replicated Z-SW, Z-SF, Z-AOV-LP, and covariate-specific Z-AOV p-values. Red vertical lines indicate conservative upper-bound p-values."
#| fig-width: 12
#| fig-height: 14
#| code-fold: true

par(mfrow = c(4, 2), mar = c(4, 4, 2.5, 2))

hist(
  sw.wbc,
  main = "Replicated Z-SW p-values: wbc model",
  breaks = 20,
  xlab = "p-value"
)
abline(v = upper_bound_pvalue(sw.wbc), col = "red", lwd = 2)

hist(
  sw.lwbc,
  main = "Replicated Z-SW p-values: lwbc model",
  breaks = 20,
  xlab = "p-value"
)
abline(v = upper_bound_pvalue(sw.lwbc), col = "red", lwd = 2)

hist(
  sf.wbc,
  main = "Replicated Z-SF p-values: wbc model",
  breaks = 20,
  xlab = "p-value"
)
abline(v = upper_bound_pvalue(sf.wbc), col = "red", lwd = 2)

hist(
  sf.lwbc,
  main = "Replicated Z-SF p-values: lwbc model",
  breaks = 20,
  xlab = "p-value"
)
abline(v = upper_bound_pvalue(sf.lwbc), col = "red", lwd = 2)

hist(
  aov.wbc.lp,
  main = "Replicated Z-AOV-LP p-values: wbc model",
  breaks = 20,
  xlab = "p-value"
)
abline(v = upper_bound_pvalue(aov.wbc.lp), col = "red", lwd = 2)

hist(
  aov.lwbc.lp,
  main = "Replicated Z-AOV-LP p-values: lwbc model",
  breaks = 20,
  xlab = "p-value"
)
abline(v = upper_bound_pvalue(aov.lwbc.lp), col = "red", lwd = 2)

hist(
  aov.wbc,
  main = "Replicated Z-AOV-wbc p-values: wbc model",
  breaks = 20,
  xlab = "p-value"
)
abline(v = upper_bound_pvalue(aov.wbc), col = "red", lwd = 2)

hist(
  aov.lwbc,
  main = "Replicated Z-AOV-log(wbc) p-values: lwbc model",
  breaks = 20,
  xlab = "p-value"
)
abline(v = upper_bound_pvalue(aov.lwbc), col = "red", lwd = 2)
```

For direct comparison, the upper-bound p-values represented by the red
vertical lines in the histograms are summarized below.

```{r table-upper-bound-pvalues}
#| code-fold: false

upper_bound_summary <- data.frame(
  Diagnostic = c(
    "Z-SW",
    "Z-SF",
    "Z-AOV-LP",
    "Z-AOV-covariate"
  ),
  `wbc model` = c(
    upper_bound_pvalue(sw.wbc),
    upper_bound_pvalue(sf.wbc),
    upper_bound_pvalue(aov.wbc.lp),
    upper_bound_pvalue(aov.wbc)
  ),
  `lwbc model` = c(
    upper_bound_pvalue(sw.lwbc),
    upper_bound_pvalue(sf.lwbc),
    upper_bound_pvalue(aov.lwbc.lp),
    upper_bound_pvalue(aov.lwbc)
  ),
  check.names = FALSE
)

knitr::kable(
  upper_bound_summary,
  digits = 6,
  align = c("l", "r", "r"),
  caption = paste(
    "Table 4: Conservative upper-bound p-values corresponding to",
    "the red vertical lines in the replicated p-value histograms."
  )
)
```

## Other Residual Calculations

### Censored Z-residuals

The normality of censored Z-residuals can be tested by an extended
Shapiro-Francia method for censored observations, implemented by
`gof.censored.zresidual()`.

```{r censored-zresiduals}
#| code-fold: false

censored.Zresid.LeukSurv.wbc <- surv_residuals(
  fit.object = fit_LeukSurv_wbc,
  data = LeukSurv,
  residual.type = "censored Z-residual"
)

censored.Zresid.LeukSurv.logwbc <- surv_residuals(
  fit.object = fit_LeukSurv_logwbc,
  data = LeukSurv,
  residual.type = "censored Z-residual"
)

gof.censored.zresidual(
  censored.Zresidual = censored.Zresid.LeukSurv.wbc
)

gof.censored.zresidual(
  censored.Zresidual = censored.Zresid.LeukSurv.logwbc
)
```

### Cox-Snell residuals

The overall GOF tests and graphical checks based on Cox-Snell residuals
suggest that both models fit the dataset adequately. For each model, the
estimated cumulative hazard function of the Cox-Snell residuals lies close to
the 45-degree reference line.

```{r fig-cox-snell}
#| fig-cap: "Cox-Snell residuals for the wbc model (left) and lwbc model (right)."
#| fig-subcap:
#|   - "wbc model"
#|   - "lwbc model"
#| layout-ncol: 2
#| fig-width: 6
#| fig-height: 5
#| code-fold: false

ucs.LeukSurv.wbc <- surv_residuals(
  fit.object = fit_LeukSurv_wbc,
  data = LeukSurv,
  residual.type = "Cox-Snell"
)

ucs.LeukSurv.logwbc <- surv_residuals(
  fit.object = fit_LeukSurv_logwbc,
  data = LeukSurv,
  residual.type = "Cox-Snell"
)

plot(
  ucs.LeukSurv.wbc,
  main.title = "Cox-Snell residuals: wbc model",
  outlier.return = FALSE
)

plot(
  ucs.LeukSurv.logwbc,
  main.title = "Cox-Snell residuals: lwbc model",
  outlier.return = FALSE
)
```

### Martingale residuals

The martingale residuals are mostly in the interval $(-3,1)$ for both models.
For the wbc model, the LOWESS curve has a slight upward slope on the left. For
the lwbc model, the curve has a more pronounced downward shape. Neither curve
is completely horizontal.

```{r fig-martingale}
#| fig-cap: "Martingale residuals for the wbc model (left) and lwbc model (right)."
#| fig-subcap:
#|   - "wbc model"
#|   - "lwbc model"
#| layout-ncol: 2
#| fig-width: 6
#| fig-height: 5
#| code-fold: false

martg.LeukSurv.wbc <- surv_residuals(
  fit.object = fit_LeukSurv_wbc,
  data = LeukSurv,
  residual.type = "martingale"
)

martg.LeukSurv.logwbc <- surv_residuals(
  fit.object = fit_LeukSurv_logwbc,
  data = LeukSurv,
  residual.type = "martingale"
)

plot(
  martg.LeukSurv.wbc,
  x_axis_var = "wbc",
  main.title = "Martingale residuals: wbc model",
  outlier.return = FALSE
)

plot(
  martg.LeukSurv.logwbc,
  x_axis_var = "logwbc",
  main.title = "Martingale residuals: lwbc model",
  outlier.return = FALSE
)
```

### Deviance residuals

The deviance residuals are more symmetric than the martingale residuals and
mostly lie in the interval $(-3,3)$. In both models, the LOWESS curves show
noticeably non-horizontal trends.

```{r fig-deviance}
#| fig-cap: "Deviance residuals for the wbc model (left) and lwbc model (right)."
#| fig-subcap:
#|   - "wbc model"
#|   - "lwbc model"
#| layout-ncol: 2
#| fig-width: 6
#| fig-height: 5
#| code-fold: false

dev.LeukSurv.wbc <- surv_residuals(
  fit.object = fit_LeukSurv_wbc,
  data = LeukSurv,
  residual.type = "deviance"
)

dev.LeukSurv.logwbc <- surv_residuals(
  fit.object = fit_LeukSurv_logwbc,
  data = LeukSurv,
  residual.type = "deviance"
)

plot(
  dev.LeukSurv.wbc,
  x_axis_var = "wbc",
  main.title = "Deviance residuals: wbc model",
  outlier.return = FALSE
)

plot(
  dev.LeukSurv.logwbc,
  x_axis_var = "logwbc",
  main.title = "Deviance residuals: lwbc model",
  outlier.return = FALSE
)
```
## References

Li, L., Wu, T., & Feng, C. (2021). Model diagnostics for censored regression
via randomized survival probabilities. *Statistics in Medicine*.
[https://doi.org/10.1002/sim.8852](https://doi.org/10.1002/sim.8852);
[publisher page](https://onlinelibrary.wiley.com/share/author/F8DKBTX7IT7UT2WTSZP3?target=10.1002/sim.8852)

Wu, T., Li, L., & Feng, C. (2023). Z-residual diagnostics for detecting
misspecification of the functional form of covariates for shared frailty
models. *arXiv*, 2302.09106.
[https://arxiv.org/abs/2302.09106](https://arxiv.org/abs/2302.09106)
